---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Parameterizing the residuals

```{r echo=FALSE}
source("libs/Common.R")
```

```{r echo = FALSE}
pkg_ver(c("ggplot2"))
```

-----------------------------------

In the univariate section of this course, we learned that parameterizing residuals offers many benefits, including the ability to succinctly characterize spread using a single parameter, such as the standard deviation. While the standard deviation is broadly applicable, it is often strongly associated with the Normal distribution. This association arises because, in many statistical methods and contexts (e.g., hypothesis testing and confidence intervals), the assumption of normality simplifies interpretation and analysis.

### Checking residuals for normality

If you are interested in conducting a hypothesis test (i.e. addressing the question *"is the slope significantly different from 0"*) you will likely want to check the residuals for normality since this is an assumption made when computing a confidence interval and a p-value. 

We learned that a Normal q-q was well suited for this comparison. For example, using the `eda_theo` function, we can compare the model's residuals to a unit Normal distribution.

```{r fig.height=2.5, fig.width=2.5}
library(tukeyedar)

M <- lm(mpg ~ hp, mtcars)
residuals <- M$residuals
eda_theo(residuals)
```

The same theoretical Q-Q plot cab be generated with `ggplot2` using the `stat_qq` and `stat_qq_line` functions.

```{r fig.height=2.5, fig.width=2.5}
ggplot() + aes(sample = residuals) + 
  stat_qq(distribution = qnorm) +
  stat_qq_line(distribution = qnorm, col = "blue")
```



## Re-expressing the data

The monotone spread can be problematic if we are to characterize the spread of `cp.ratio` as being the same (homogeneous) across the full range of `area` values. To remedy this, we can re-express the `cp.ratio` values. Variables measured as ratios (such is the case with `cp.ratio`) are good candidates for log transformation. We will therefore fit a new linear model to the data after transforming the y-value. 

```{r}
df.log <- data.frame( area = df$area, cp.ratio.log = log(df$cp.ratio))
M3     <- lm(cp.ratio.log ~ area, dat = df.log)
```

Transforming a variable, whether $x$ or $y$, will change the nature of the relationship between both variables. We therefore need to recreate the scatter plot. We'll also see how well a 1^st^ order polynomial can characterize this relationship (re-expressing values can sometimes help "straighten" a relationship between variables, hence why we are not starting off first with a 2^nd^ order polynomial fit).

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
ggplot(df.log, aes(x = area, y = cp.ratio.log)) + geom_point() +
  stat_smooth(method = "lm", se = FALSE)
```
 
At first glance, the log transformation seems to have done a good job at straightening the batch of values. Next, let's look at the residual-dependence plot.

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
df.log$residuals <- residuals(M3)
ggplot(df.log, aes(x = area, y = residuals)) + geom_point() +
               stat_smooth(method = "loess", se = FALSE, span = 1, 
                           method.args = list(degree = 1) )
```

Logging the y values has eliminated the residual's dependence on `area`. Next, let's assess homogeneity in the residuals using the s-l plot.

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
sl3 <- data.frame( std.res = sqrt(abs(residuals(M3))), 
                   fit     = predict(M3))
ggplot(sl3, aes(x = fit, y = std.res)) + geom_point() +
               stat_smooth(method ="loess", se = FALSE, span = 1, 
                           method.args=list(degree = 1) )
```

Even though we observe a mid range peak in spread, we do not observe a **systematic increase** in spread. The log transformation seems to have removed the monotone spread. 

Finally, we'll check for normality of the residuals.

```{r fig.height=2.5, fig.width=2.5}
ggplot(df.log, aes(sample = residuals)) + 
  geom_qq(distribution = qnorm) +
  geom_qq_line(distribution = qnorm, col = "blue")
```

Re-expressing the data seems to have improved the *normality* of the residuals somewhat. Though not perfect, more points seem to follow a straight line. But, recall from chapter 18.4 that noise can be quite prominent in a normal q-q plot when working with small datasets.

[^1]: Cleveland, William. *Visualizing data*. Hobart Press. 1993. [pp 87-88]
