---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Visualizing Variability Decomposition in Bivariate Models

In bivariate analysis, a key objective is to identify a model that maximizes our ability to explain the variability in the response variable, Y. The more variability in Y that is explained by the model, the smaller the uncertainty in estimating Y for each value of the predictor variable, X. This goal translates into minimizing the spread of residuals relative to the variability captured by the model.

The figure below illustrates this concept using three scenarios of increasing model complexity:

+ **Left plot**: A **mean-only model** (0^th^-order polynomial) predicts the same constant value for all observations. The red points represent the predicted values, and the vertical dashed lines show the residuals—the differences between the observed (gray) and predicted (red) values. Since the model does not account for any variation in  X, all variability in Y remains unexplained.

+ **Middle plot**: A **linear model** (1^st^-order polynomial) is fitted to the data. The predicted values now vary with X, and the residuals are smaller than in the mean-only model. The model explains some of the variability in Y, as reflected in the increased spread of the red points.

+ **Right plot**: A **better-fitting linear model** (also a 1^st^-order polynomial) explains a larger portion of the variability in Y. The predicted values span a wider range (about 11 units), and the residuals are relatively small—averaging about 1 unit—indicating that most of the variability in Y is now accounted for by the model.


```{r echo = FALSE, results='hide', fig.width=9, fig.height=3}
library(tukeyedar)

n <- 20
set.seed(123)
x <- sort(runif(n, 1, 10))
y <- 1 + 1.5 * x + rnorm(n)
#y[20] <- 28
df <- data.frame(x,y)


OP <- par(mfrow = c(1,3), cex = 1)

# Scenario 1
M0 <- eda_lm(df, x, y, show.par = FALSE, mean.l = FALSE, sd = FALSE, poly = 0)
error_segs <- lapply(1:nrow(df),
       FUN = \(i) data.frame(x = df$x[i], y = c(M0$fitted.values[i], df$y[i]),
                             row.names = NULL))
OP2 <- par(mar = c(0,0,0,0))
  lapply(error_segs, \(x) lines(x, col = "grey", lty = 2))
  points(x = df$x, y = M0$fitted.values, pch = 16, col = "darkred")
par(OP2)

# Scenario 2
int1 <- 6.5
slp1 <- .5
fitted_manual <- int1 + slp1 * x
residuals_manual <- y - fitted_manual
data.frame(x, y, fitted_manual, residuals_manual)
eda_lm(df, x, y, show.par = FALSE, mean.l = FALSE, sd = FALSE, reg = FALSE)
abline(a=int1, b=slp1, col = "red")
error_segs <- lapply(1:nrow(df),
       FUN = \(i) data.frame(x = df$x[i], y = c(fitted_manual[i], df$y[i]),
                             row.names = NULL))
OP2 <- par(mar = c(0,0,0,0))
  lapply(error_segs, \(x) lines(x, col = "grey", lty = 2))
  points(x = df$x, y = fitted_manual, pch = 16, col = "darkred")
par(OP2)


# Scenario 3
M1 <- eda_lm(df, x, y, show.par = FALSE, mean.l = FALSE, sd = FALSE)
error_segs <- lapply(1:nrow(df),
       FUN = \(i) data.frame(x = df$x[i], y = c(M1$fitted.values[i], df$y[i]),
                             row.names = NULL))
OP2 <- par(mar = c(0,0,0,0))
  lapply(error_segs, \(x) lines(x, col = "grey", lty = 2))
  points(x = df$x, y = M1$fitted.values, pch = 16, col = "darkred")
par(OP2)

par(OP)
```

While scatter plots like these can provide insight into model performance—especially in extreme cases—they are not always ideal for assessing how well a model explains variability across the full range of the data. A **variability decomposition (VD)** plot, first introduced in Chapter 21, is specifically designed to evaluate this aspect. It provides a more structured and comparative view of how much of the total variability is captured by the model versus left in the residuals.

In the next section, we’ll use the variability decomposition plot to formalize and visualize these differences more clearly.


## Conceptual Overview

The variability decomposition plot visually partitions the total spread of the response variable into:

+ **Fitted values**: representing the portion of variability explained by the model.
+ **Residuals**: representing the portion of variability not explained by the model.

This decomposition is analogous to the concept behind $R^2$, but it is presented graphically to support exploratory analysis and intuitive understanding.


## Model Progression Examples

To illustrate how a model *fit* affects variability decomposition, we use synthetic data where the relationship between $X$ and $Y$ is controlled. This allows us to showcase three scenarios:

### Scenario 1: 0th-Order Polynomial (Mean-Only Model)

In this case, the model fits only the mean of \( Y \), ignoring any relationship with \( X \). All variability in \( Y \) is captured by the residuals.

```{r fig.width=9, fig.height=3, echo = FALSE}
set.seed(321)
n <- 100
x <- sort(runif(n, 1, 10))
y <- 1 + 1.5 * x + 0.5 * x^2 +rnorm(n)
df <- data.frame(x,y)

OP <- par(mfrow = c(1,3), cex = 1)
M0 <- eda_lm(df, x, y, mean.l = FALSE, sd = FALSE, poly = 0, show.par = FALSE)
plot(M1, ylim = c(-20,40), asp = FALSE )
eda_vd(M0, type = "box")
par(OP)

```


### Scenario 2: 1st-Order Polynomial (Mean-Only Model)

Here, a linear relationship between $X$ and $Y$ is introduced. The model captures quite a bit of the variability, and the residuals account for the rest.

```{r fig.width=9, fig.height=3, echo = FALSE}
OP <- par(mfrow = c(1,3))
M1 <- eda_lm(df, x, y, mean.l = FALSE, sd = FALSE, poly = 1)
plot(M1, ylim = c(-20,40), asp = FALSE )
eda_vd(M1, type = "box")
par(OP)
```

### Scenario 3: 2nd-Order Polynomial (Mean-Only Model)

In this case, the model fits only the mean of $Y$, ignoring any relationship with \( X \). All variability in \( Y \) is captured by the residuals.

```{r fig.width=6, fig.height=3, echo = FALSE}
OP <- par(mfrow = c(1,2))
M2 <- eda_lm(df, x, y, mean.l = FALSE, sd = FALSE, poly = 2)
eda_vd(M2, type = "box")
par(OP)
```

```{r fig.width=7, fig.height=3, echo = FALSE}
OP <- par(mfrow = c(1,3))
eda_vd(M0, type = "box", lim = c(-30,40))
eda_vd(M1, type = "box", lim = c(-30,40))
eda_vd(M2, type = "box", lim = c(-30,40))
par(OP)
```

## Generating a variability decomposition plot

The variability decomposition  (VD) plot can be generated with the `eda_vd` function from the `tukeyedar` package. It takes as input a model generated by `lm`.

In an earlier chapter, it was determined that a 2^nd^ order polynomial did a good job in capturing the relationship between `mpg` and `hp` from the `mtcars` dataset. We will therefore work off of that model in this example. A side-by-side plot of the fitted model and the resulting variability decomposition plots are shown next.

```{r class.source="eda", fig.width = 6, fig.height= 3.2, eval=FALSE}
library(tukeyedar)

# Fit the model to the data
M2 <- lm(mpg ~ hp + I(hp^2), mtcars)
plot(mpg ~ hp, mtcars)
x.pred <- data.frame( hp = seq(min(mtcars$hp), max(mtcars$hp), length.out = 50))
y.pred <- predict(M2, x.pred)
lines(x.pred$hp, y.pred, col = "red")

# Generate the variability decomposition plot
eda_vd(M2, type = "box")
```

```{r class.source="eda", fig.width = 6, fig.height= 3.2, echo=FALSE}
OP <- par(mfrow = c(1,2))
library(tukeyedar)

# Fit the model to the data
M2 <- lm(mpg ~ hp + I(hp^2), mtcars)
OP2 <- par(mar = c(3,3,3,1), col = "grey30", col.lab = "grey40" , col.axis = "grey40")
plot(mpg ~ hp, mtcars, xlab = "", ylab = "")
mtext("hp", side = 1, line = 2)  # default is around 2.5-3
mtext("mpg", side = 2, line = 2)
x.pred <- data.frame( hp = seq(min(mtcars$hp), max(mtcars$hp), length.out = 50))
y.pred <- predict(M2, x.pred)
lines(x.pred$hp, y.pred, col = "red")
par(OP2)
# Generate the variability decomposition plot
eda_vd(M2, type = "box")
par(OP)
```

