---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Dependence of $y$ on $x$

```{r echo=FALSE}
source("libs/Common.R")
```

```{r echo = FALSE}
pkg_ver(c("ggplot2"))
```

---------------------------------

Bivariate data consist of datasets where *two* variables are measured for each observation. For instance, both wind speed *and* temperature might be recorded at specific locations. This contrasts with univariate data, where only a *single* variable—such as temperature—is measured for each observation at *two* different locations.

Understanding how to model and analyze bivariate data is critical for uncovering relationships and trends between the variables. In this chapter, we will delve into common fitting strategies that are specifically tailored for bivariate datasets.

## Scatter plot

A scatter plot is a widely used visualization tool for comparing values between two variables. In some cases, one variable is considered *dependent* on the other, which is then referred to as the *independent* variable. The dependent variable, also known as the *response* variable, is typically plotted on the y-axis, while the independent variable, also called the *factor* (not to be confused with the factor data type in R, used as a grouping variable), is plotted on the x-axis.

In other situations, the focus may not be on a dependent-independent relationship. Instead, the goal might simply be to explore and understand the overall relationship between the two variables, without assigning a hierarchical dependency.

The following figure shows a scatter plot of a vehicle's miles-per-gallon (mpg) consumption as a function of horsepower (hp). In this exercise, we'll seek to understand how a car's miles-per-gallon (the dependent variable) can be related to its horsepower (the independent variable).

```{r class.source="eda", fig.height = 3, fig.width = 3, echo = FALSE}
library(tukeyedar)
eda_lm(mtcars, hp, mpg, lm = FALSE, sd= FALSE, show.par = FALSE, mean.l = FALSE)
```

## Fitting a model to the data

In univariate data analysis, we aim to reduce the data to manageable parameters, providing a clearer "handle" on the dataset. Similarly, when working with bivariate data, we can start by fitting the simplest possible model. For the variable `mpg`, a straightforward approach is to use a measure of location, such as the mean.

```{r class.source="eda", fig.height = 3, fig.width = 3, echo = FALSE}
eda_lm(mtcars, hp, mpg, poly=0, sd= FALSE, show.par = FALSE, mean.l = FALSE)
```

The red line represents the fitted model.  Mathematically, this model can be expressed as:

$$
mpg = a + b(hp)^0 = a + b(1) + \epsilon = 20.09 + \epsilon
$$
The equation is a 0^th^ order polynomial model, The polynomial order is determined by the highest power to which the independent variable is raised (0 in this case). For now, note that regardless of `hp`'s observed value, its contribution to `mpg` remains constant. The terms $a$ and $b$ are defined as the model's **intercept** and **slope**. We will delve into the meanings of $a$ and $b$ later in this section. $\epsilon$ is the residual or the difference between each observation and the fitted model. 

If this were a univariate dataset, modeling the data using the mean would be a reasonable choice. But, given that this is a bivariate dataset, this approach fails to utilize the variation in `hp` to explain `mpg`.

The pattern observed in the scatter plot suggests that as `hp` increases, `mpg` decreases. We can model this by fitting the line in such a way so as to capture the pattern generated by the points. The fitting strategy can be as simple as "eyeballing" the fit, or adopting more involved statistical methods like the least-squares method.

The following plot is an example of a line fitted to the data using the least-square method.

```{r class.source="eda", fig.height = 3, fig.width = 3, echo = FALSE}
eda_lm(mtcars, hp, mpg, poly=1, sd= FALSE, show.par = FALSE, mean.l = FALSE)
```

The modeled line can be expressed mathematically as:

$$
mpg = a + b(hp)^1 + \epsilon = 30.1 - 0.068(hp) + \epsilon
$$

This is a 1^st^ order polynomial where `hp` is raised to the power of 1. Here, the intercept $a$ is the value `mpg` would take if `hp` were zero. The term $b$ is the slope of the line and is interpreted as the change in $y$ for every unit change in $x$. In our example, for every one horsepower increase, the miles-per-gallon decreases by 0.068 miles-per-gallon.

Not all relationships will necessarily follow a perfectly straight line. Relationships can be curvilinear. Polynomial functions can take on many different non-linear shapes. The power to which the $x$ variable is raised to will define the nature of that shape. For example, to fit a quadratic line to the data, one can define the following model:

$$
mpg = a + b(hp) + c(hp)^2 + \epsilon 
$$

The model retains the *slope* term, $b$, and adds the *curvature coefficient*, $c$.

Fitting the model to the data using least squares method results in the following coefficients:

$$
mpg = 40.4 -0.2(hp) + 0.0004(hp)^2 + \epsilon 
$$

The modeled line looks like:

```{r class.source="eda", fig.height = 3, fig.width = 3, echo = FALSE}
eda_lm(mtcars, hp, mpg, poly=2, sd= FALSE, show.par = FALSE, mean.l = FALSE)
```

## Generating bivariate plots with base R

In R, a scatter plot can be easily created using the base plotting environment. Below, is an example of this process with the built-in `mtcars` dataset.

```{r fig.width=2.5, fig.height=2.5, small.mar=TRUE}
plot(mpg ~ hp, mtcars)
```

The formula `mpg ~ hp` can be interpreted as *...mpg as a function of hp...*, where `mpg` (miles per gallon) is plotted on the y-axis and `hp` (horsepower) on the x-axis.

To overlay a regression line on the scatter plot, you first need to define a linear model using the `lm()` function.

```{r}
M1 <- lm(mpg ~ hp, mtcars)
```

Here, a first-order polynomial (linear) model is fitted to the data. The `mpg ~ hp` carries the same interpretation as in the `plot()` function. The model coefficients (intercept and slope) can be extracted as a numeric vector from the model `M1` using the `coef()` function.

```{r}
coef(M1)
```

Once the model is defined, you can add the regression line to the plot by passing the `M1` model to  `abline()` as an argument.

```{r fig.width=2.5, fig.height=2.5, small.mar=TRUE}
plot(mpg ~ hp, mtcars)
abline(M1, col = "red")
```

In this example, the regression line is displayed in red.

To fit a second-order polynomial (quadratic) model, extend the formula by adding a second-order term using the `I()` function:

```{r}
M2 <- lm(mpg ~ hp + I(hp^2), mtcars)
```

When raising an independent variable $x$ to a power using the caret (`^`) operator, you must wrap the term with `I()`. Without it, `lm()` may misinterpret the formula, leading to unintended behavior. 

Adding the quadratic regression line (or any higher-order polynomial model) to the plot requires generating predicted values using the `predict()` function. These predictions are then plotted using the `lines()` function:

```{r fig.width=2.5, fig.height=2.5, small.mar=TRUE}
plot(mpg ~ hp, mtcars)

# Create sequence of hp values to 
x.pred <- data.frame( hp = seq(min(mtcars$hp), max(mtcars$hp), length.out = 50))

# Predict mpg values based on the quadratic model
y.pred <- predict(M2, x.pred)

# Add the quadratic regression line to the plot
lines(x.pred$hp, y.pred, col = "red")
```

In this code, we:

+  Create a sequence of `hp` values covering the range of the dataset.
+  Predict the corresponding `mpg` values using the quadratic model `M2`.
+  Add the predicted curve to the plot with `lines()`.

## Generating bivariate plots with `ggplot`

Using the `ggplot2` package in R, a scatter plot can be created with minimal effort. Here's an example using the built-in `mtcars` dataset:

```{r fig.width=2.5, fig.height=2.5}
library(ggplot2)

ggplot(mtcars, aes(x = hp, y = mpg)) + geom_point()
```

In this code, `ggplot()` initializes the plot and specifies the dataset (`mtcars`) and aesthetics (`aes()`), where x = hp and y = mpg. `geom_point()` adds points to create the scatter plot.

o add a linear regression line to the scatter plot, use the `stat_smooth()` function:

```{r fig.width=2.5, fig.height=2.5}
ggplot(mtcars, aes(x = hp, y = mpg)) + geom_point() + 
             stat_smooth(method ="lm", se = FALSE)
```

Here, the `method = "lm"` argument specifies that a linear model (`lm`) is used for the regression line. The `se = FALSE` argument prevents the addition of a confidence interval around the regression line. (Confidence intervals are not covered in this course). Note that there’s no need to create a separate model outside the ggplot pipeline, as the `stat_smooth()` function handles it automatically.

The `stat_smooth()` function can also fit higher-order polynomial models by specifying the desired formula. For instance, to fit a quadratic model:

```{r fig.width=2.5, fig.height=2.5}
ggplot(mtcars, aes(x = hp, y = mpg)) + geom_point() + 
             stat_smooth(method ="lm", se = FALSE, y = formula = y ~  x + I(x^2) )
```

In this case, the formula argument defines the model as $y=x+x^2$, where $x$ and $y$ refer to the x- and y-axes, respectively. The `I()` function ensures that the quadratic term (`x^2`) is treated correctly in the formula.

It’s worth noting that the formula directly references `x` and `y` instead of specific column names. This generality allows the formula to adapt to the axis mappings defined in the `aes()` function.

## None-parametric fit

Polynomial models used to fit lines to data are classified as **parametric models**. These models require defining a specific functional form (e.g., linear, quadratic) *a priori*, which is then fitted to the data. By imposing this predefined structure, parametric models make strong assumptions about the underlying relationship between variables.

In contrast, **non-parametric models** belong to a class of fitting strategies that do not assume a specific structure for the data. Instead, these models are designed to adapt flexibly, allowing the data to reveal its inherent patterns and relationships. One such method used in this course is the **loess** fit, a locally weighted regression approach.

#### Loess

A flexible curve-fitting method is the **loess**  curve (short for **lo**cal regr**ess**ion, also known as *local weighted regression*). This technique fits small segments of a regression line across the range of x-values and links the midpoints of these segments to generate a *smooth* curve. The range of x-values contributing to each localized regression lines is controlled by the **span** parameter,  $\alpha$,  which typically ranges from 0.2 to 1 (though it can exceed 1 for smaller datasets). A larger $\alpha$ value results in a smoother curve. Another key parameter, $\lambda$, specifies the **polynomial order** of the localized regression lines. This is usually set to 1, although in `ggplot2`, the loess function defaults to a 2^nd^ order polynomial.

```{r echo=FALSE}
library(dplyr)
library(purrr)

#df <- read.csv("http://mgimond.github.io/ES218/Data/ganglion.csv")
df <- mtcars
alpha <-  0.5  
strip.x <- nrow(df) * alpha # Number of points within band

f.plot <- function(start , line = FALSE, point = FALSE, bnds =FALSE,
                   w = FALSE, plot = TRUE, pts = FALSE, title = NULL){
  # Find points closest to starting point
  subset <-  df %>% 
  #  mutate(dst = abs(area - start)) %>% 
    mutate(dst = abs(hp - start)) %>%
    arrange(dst) %>% 
    mutate(j = row_number()) %>% 
    filter(j <= strip.x ) %>% 
    mutate(wt = dst / max(dst) * 3)
  
  # Assign weights
  wts <- dnorm(subset$wt)/ 0.3989423
  
  # Regress with weights
#  M <- lm(cp.ratio ~ area, subset, weights = wts)
  M <- lm(mpg ~ hp, subset, weights = wts)
  x.l <- coef(M)[1] + coef(M)[2] * start
  
  # Plot by option
  if(plot == TRUE){
 #   plot(cp.ratio ~ area, df,  yaxt='n', main = title,
    plot(mpg ~ hp, df,  yaxt='n', main = title,
         axes = FALSE, pch=16, col = "grey90", cex = 1.6)
  #  axis(side=1, at=c(seq(10,150,30)))
    axis(side=1, at=c(seq(10,350,30)))
    abline(v = start, lty = 2)
    if(bnds == TRUE){
   #   abline(v = c(min(subset$area),max(subset$area)), lty = 3, col = "grey")
   #   rect(min(subset$area), 0, max(subset$area), 20, col = rgb(0,0,1,0.1),
       abline(v = c(min(subset$hp),max(subset$hp)), lty = 3, col = "grey")
   #    rect(min(subset$hp), 0, max(subset$hp), 20, col = rgb(0,0,1,0.1),
       rect(min(subset$hp), 0, max(subset$hp), 35, col = rgb(0,0,1,0.1),      
           border = rgb(0,0,1,0.1))
    }
    
    if(w == TRUE){
    #  points(x = subset$area, y = subset$cp.ratio, col = rgb(0,0,1, wts), 
       points(x = subset$hp, y = subset$mpg, col = rgb(0,0,1, wts),
             pch = 16,  cex=1.6)
    }
    if(line == TRUE){
      # clip(min(subset$area),max(subset$area),
      #      min(df$cp.ratio),max(df$cp.ratio))
      clip(min(subset$hp),max(subset$hp),
           min(df$mpg),max(df$mpg))
      abline(M, col = "orange", lwd = 1.8)
    }
    if(point == TRUE){
      points(x = start, y = x.l, pch = 16, col = "red", cex=1.8) 
    }
  }
  
  if(pts == TRUE){
    return(c(start, x.l))
  }  
  
}

```

#### How a loess is constructed

Behind the scenes, each point ($x_i$, $y_i$) that defines the loess curve is constructed as follows: 
 
 a) A subset of data points closest to point $x_i$ are identified ($x_{130}$ is used as an example in the figure below). The number of points in the subset is identified by multiplying the bandwidth $\alpha$ by the total number of observations. In our current example, $\alpha$ is set to `r alpha`. The number of points defining the subset is thus 0.5 * 32 = 16. The region encompassing these points is displayed in a light blue color.
 
 b) The points in the subset are assigned weights. Greater weight is assigned to points closest to $x_i$. The weights define the points' influence on the fitted line. Different weighting techniques can be implemented in a loess with the `gaussian` weight being the most common. Another weighting strategy we will also explore later in this course is the `symmetric` weight.
 
 c) A regression line is fit to the subset of points. Points with smaller weights will have less leverage on the fitted line than points with larger weights. The fitted line can be either a first order polynomial fit or a second order polynomial fit.
 
 d) Next, the value $y_i$ from the regression line is computed (red point in panel (d)). This is one of the points that will define the shape of the loess.
   

```{r fig.height = 2.7, fig.width = 10, echo = FALSE}
# OP <- par(mfrow = c(1, 4), mar=c(2,1,1,0), pty = "s")
# f.plot(start = 50, bnds = TRUE, title = "(a) Subset points") 
# f.plot(start = 50, bnds = TRUE, w = TRUE, title = "(b) Assign weights")
# f.plot(start = 50, bnds = TRUE, w = TRUE, line = TRUE, title = "(c) Fit line") 
# f.plot(start = 50,  bnds = TRUE, w = TRUE, line = TRUE, point = TRUE, 
#        title = "(d) Draw point")
# par(OP)

OP <- par(mfrow = c(1, 4), mar=c(2,1,1,0), pty = "s")
f.plot(start = 130, bnds = TRUE, title = "(a) Subset points") 
f.plot(start = 130, bnds = TRUE, w = TRUE, title = "(b) Assign weights")
f.plot(start = 130, bnds = TRUE, w = TRUE, line = TRUE, title = "(c) Fit line") 
f.plot(start = 130,  bnds = TRUE, w = TRUE, line = TRUE, point = TRUE, 
       title = "(d) Draw point")
par(OP)
```

The above steps are repeated for as many $x_i$ values practically possible. Note that when $x_i$ approaches an upper or lower limit, the subset of points is no longer centered on $x_i$. For example, when estimating $x_{52}$, the sixteen closest points to the right of $x_{52}$ are selected. Likewise, for the upper bound $x_{335}$, the sixteen closest points to the left of  $x_{335}$ are selected.

```{r fig.height = 2.5, fig.width = 10, echo = FALSE}
# OP <- par(mfrow = c(1, 4), mar=c(2,1,1,0), pty = "s")
# f.plot(start = 10,  bnds = TRUE, w = TRUE, line = TRUE, point = TRUE, 
#        title = expression(Left-most ~~  x[i] )) 
# f.plot(start = 140,  bnds = TRUE, w = TRUE, line = TRUE, point = TRUE, 
#        title = expression(Right-most ~~  x[i]) ) 
# par(OP)

OP <- par(mfrow = c(1, 4), mar=c(2,1,1,0), pty = "s")
f.plot(start = 52,  bnds = TRUE, w = TRUE, line = TRUE, point = TRUE, 
       title = expression(Left-most ~~  x[i] )) 
f.plot(start = 335,  bnds = TRUE, w = TRUE, line = TRUE, point = TRUE, 
       title = expression(Right-most ~~  x[i]) ) 
par(OP)
```


In the following example, just under 30 loess points are computed at equal intervals. This defines the shape of the loess.

```{r fig.height = 2.3, fig.width = 10, echo = FALSE}
# OP <- par(mfrow = c(1, 4), mar=c(2,1,1,0), pty = "s")
# l.pts <- seq(10,140,(110-20)/20) %>% map(function(x) f.plot(start = x, plot = FALSE, pts = TRUE)) %>% 
#   do.call(rbind, .) %>% as.data.frame()
# plot(cp.ratio ~ area, df,  yaxt='n', main = NULL,
#          axes = FALSE, pch=16, col = "grey90", cex = 1.6)
#     axis(side=1, at=c(seq(10,150,30)))
# points(l.pts$V1, l.pts$`(Intercept)`, pch = 16, col = "red",
#        cex=1)
# par(OP)

OP <- par(mfrow = c(1, 4), mar=c(2,1,1,0), pty = "s")
l.pts <- seq(50,330,(330-50)/20) %>% map(function(x) f.plot(start = x, plot = FALSE, pts = TRUE)) %>% 
  do.call(rbind, .) %>% as.data.frame()
plot(mpg ~ hp, df,  yaxt='n', main = NULL,
         axes = FALSE, pch=16, col = "grey90", cex = 1.6)
    axis(side=1, at=c(seq(50,335,30)))
points(l.pts$V1, l.pts$`(Intercept)`, pch = 16, col = "red",
       cex=1)
par(OP)
```

It's more conventional to plot the line segments than it is to plot the points.

```{r fig.height = 2.3, fig.width = 10, echo = FALSE}
# OP <- par(mfrow = c(1, 4), mar=c(2,1,1,0), pty = "s")
# l.pts <- seq(10,140,(110-20)/20) %>% map(function(x) f.plot(start = x, plot = FALSE, pts = TRUE)) %>% 
#   do.call(rbind, .) %>% as.data.frame()
# plot(cp.ratio ~ area, df,  yaxt='n', main = NULL,
#          axes = FALSE, pch=16, col = "grey90", cex = 1.6)
#     axis(side=1, at=c(seq(10,150,30)))
# lines(l.pts$V1, l.pts$`(Intercept)`, col = "red")
# par(OP)

OP <- par(mfrow = c(1, 4), mar=c(2,1,1,0), pty = "s")
l.pts <- seq(50,335,(335-50)/20) %>% map(function(x) f.plot(start = x, plot = FALSE, pts = TRUE)) %>% 
  do.call(rbind, .) %>% as.data.frame()
plot(mpg ~ hp, df,  yaxt='n', main = NULL,
         axes = FALSE, pch=16, col = "grey90", cex = 1.6)
    axis(side=1, at=c(seq(50,335,30)))
lines(l.pts$V1, l.pts$`(Intercept)`, col = "red")
par(OP)

```

<br>

## Generating a loess model in base R

The loess fit can be computed in R using the `loess()` function. It takes as arguments  `span` ($\alpha$), and `degree` ($\lambda$).

```{r}
# Fit loess function
lo <- loess(mpg ~ hp, mtcars, span = 0.5, degree = 1)

# Predict loess values for a range of x-values
lo.x <- seq(min(mtcars$hp), max(mtcars$hp), length.out = 50)
lo.y <- predict(lo, lo.x)
```

The modeled loess curve can be added to the scatter plot using the `lines` function.

```{r fig.width=2.5, fig.height=2.5, small.mar=TRUE}
plot(mpg ~ hp, mtcars)
lines(lo.x, lo.y, col = "red")
```

## Generating a loess model in `ggplot`

In `ggplot2` simply pass the `method="loess"` parameter to the `stat_smooth` function.

```{r eval=FALSE}
ggplot(mtcars, aes(x = hp, y = mpg)) + geom_point() + 
             stat_smooth(method = "loess", se = FALSE, span = 0.5)
```

`ggplot` defaults to a second degree loess (i.e. the small regression line elements that define the loess are modeled using a 2^nd^ order polynomial and not a 1^st^ order polynomial). If a first order polynomial is desired, you need to include an argument list in the form of `method.args=list(degree=1)` to the `stat_smooth` function.

```{r fig.width=2.5, fig.height=2.5}
ggplot(mtcars, aes(x = hp, y = mpg)) + geom_point() + 
             stat_smooth(method = "loess", se = FALSE, span = 0.5, 
                         method.args = list(degree = 1) )
```

## Residuals

Fitting the data with a line is just the first step in EDA. Your next step should be to explore the residuals. The residuals are the distances (parallel to the y-axis) between the observed points and the fitted line. The closer the points are to the line (i.e. the smaller the residuals) the better the fit. 

The residuals can be computed using the `residuals()` function. It takes as argument the model object. For example, to extract the residuals from the linear model `M` computed earlier type, 

```{r}
residuals(M)
```

### Residual-dependence plot

One way to visualize the residuals is to create a **residual-dependence plot**--this plots the residuals as a function of the x-values. We'll do this using `ggplot` so that we can also fit a loess curve to help discern any pattern in the residuals.


```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
df$residuals <- residuals(M)
ggplot(df, aes(x = area, y = residuals)) + geom_point() +
             stat_smooth(method = "loess", se = FALSE, span = 1, 
                         method.args = list(degree = 1) )

```

We are interested in identifying any pattern in the residuals. **If the model does a good job in fitting the data, the points should be uniformly distributed across the plot** and the loess fit should approximate a horizontal line. With the linear model `M`, we observe a convex pattern in the residuals suggesting that the linear model is not a good fit. We say that the residuals show *dependence* on the x values. 

Next, we'll look at the residuals from the second order polynomial model `M2`.

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
df$residuals2 <- residuals(M2)
ggplot(df, aes(x = area, y = residuals2)) + geom_point() +
               stat_smooth(method = "loess", se = FALSE, span = 1, 
                           method.args = list(degree = 1) )

```

There is no indication of dependency between the residual and the `area` values. The second order polynomial is a large improvement over the first order polynomial. Let's look at the loess model.


```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
df$residuals3 <- residuals(lo)
ggplot(df, aes(x = area, y = residuals3)) + geom_point() +
               stat_smooth(method = "loess", se = FALSE, span = 1, 
                           method.args = list(degree = 1) )

```

No surprise here. The loess model does a good job in capturing any overall pattern in the data. But note that this is to be expected given that the loess fit is a non-parametric model--it lets the data define its shape!

> You may ask *“if the loess model does such a good job in fitting the data, why bother with polynomial fits?”* If you are seeking to generate a predictive model that explains the relationship between the y and x variables, then a mathematically tractable model (like a polynomial model) should be sought. If the interest is simply in identifying a pattern in the data, then a loess fit is a good choice.

A model is deemed "well defined" when its residuals are constant over the full range of $x$. In essence, we expect a formula  of the form:

$$
Y = a + bx + \varepsilon
$$
where $\varepsilon$ is a constant that does not vary as a function of varying $x$.  This should sound quite familiar to you given that we've spent a good part of the univariate analysis section seeking a homogeneous spread in the data (i.e. a spread that did not change as a function of fitted group values). So what other diagnostic plots can we (and *should we*) generate from the fitted model? We explore such plots next.


### Spread-location plot

The `M2` and `lo` models do a good job in eliminating any dependence between residual and x-value. Next, we will check that **the residuals do not show a dependence with *fitted* y-values**. This is analogous to univariate analysis where we checked if residuals increased or decreased with increasing medians across categories. Here we will compare residuals to the fitted `cp.ratio` values (for a univariate analogy, think of the fitted line as representing a *level* across different segments along the x-axis). We'll generate a spread-level plot of model `M2`'s residuals (note that in the realm of regression analysis, such plot is often referred to as a **scale-location** plot). We'll also add a loess curve to help visualize any patterns in the plot.

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
sl2 <- data.frame( std.res = sqrt(abs(residuals(M2))), 
                   fit     = predict(M2))

ggplot(sl2, aes(x = fit, y  =std.res)) + geom_point() +
              stat_smooth(method = "loess", se = FALSE, span = 1, 
                          method.args = list(degree = 1) )
```

The function `predict()` extracts the y-values from the fitted model `M2` and is plotted along the x-axis. It's clear from this plot that the residuals are not homogeneous; they increase as a function of increasing *fitted* CP ratio. The "bend" observed in the loess curve is most likely due to a single point at the far (right) end of the fitted range. Given that we have a small batch of numbers, a loess can be easily influenced by an outlier. We may therefore want to increase the loess span by setting `span = 2`.


```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
ggplot(sl2, aes(x = fit, y = std.res)) + geom_point() +
              stat_smooth(method = "loess", se = FALSE, span = 2, 
                          method.args = list(degree = 1) )
```

The point's influence is reduced enough to convince us that the observed monotonic increase is real. 

We learned with the univariate data that re-expressing values was one way to correct for any increasing or decreasing spreads across the range of fitted values. At this point, we may want to look into re-expressing the data. But, before we do, we'll explore another diagnostic plot: the *normal q-q plot*.

### Checking residuals for normality

If you are interested in conducting a hypothesis test (i.e. addressing the question *"is the slope significantly different from 0"*) you will likely want to check the residuals for normality since this is an assumption made when computing a confidence interval and a p-value. 

In `ggplot`, we learned that a normal q-q plot could be generated using the `stat_qq` and `stat_qq_line` function.

```{r fig.height=2.5, fig.width=2.5}
ggplot(df, aes(sample = residuals2)) + 
  stat_qq(distribution = qnorm) +
  stat_qq_line(distribution = qnorm, col = "blue")
```

Here, the residuals seem to stray a little from a normal distribution.

## Re-expressing the data

The monotone spread can be problematic if we are to characterize the spread of `cp.ratio` as being the same (homogeneous) across the full range of `area` values. To remedy this, we can re-express the `cp.ratio` values. Variables measured as ratios (such is the case with `cp.ratio`) are good candidates for log transformation. We will therefore fit a new linear model to the data after transforming the y-value. 

```{r}
df.log <- data.frame( area = df$area, cp.ratio.log = log(df$cp.ratio))
M3     <- lm(cp.ratio.log ~ area, dat = df.log)
```

Transforming a variable, whether $x$ or $y$, will change the nature of the relationship between both variables. We therefore need to recreate the scatter plot. We'll also see how well a 1^st^ order polynomial can characterize this relationship (re-expressing values can sometimes help "straighten" a relationship between variables, hence why we are not starting off first with a 2^nd^ order polynomial fit).

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
ggplot(df.log, aes(x = area, y = cp.ratio.log)) + geom_point() +
  stat_smooth(method = "lm", se = FALSE)
```
 
At first glance, the log transformation seems to have done a good job at straightening the batch of values. Next, let's look at the residual-dependence plot.

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
df.log$residuals <- residuals(M3)
ggplot(df.log, aes(x = area, y = residuals)) + geom_point() +
               stat_smooth(method = "loess", se = FALSE, span = 1, 
                           method.args = list(degree = 1) )
```

Logging the y values has eliminated the residual's dependence on `area`. Next, let's assess homogeneity in the residuals using the s-l plot.

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
sl3 <- data.frame( std.res = sqrt(abs(residuals(M3))), 
                   fit     = predict(M3))
ggplot(sl3, aes(x = fit, y = std.res)) + geom_point() +
               stat_smooth(method ="loess", se = FALSE, span = 1, 
                           method.args=list(degree = 1) )
```

Even though we observe a mid range peak in spread, we do not observe a **systematic increase** in spread. The log transformation seems to have removed the monotone spread. 

Finally, we'll check for normality of the residuals.

```{r fig.height=2.5, fig.width=2.5}
ggplot(df.log, aes(sample = residuals)) + 
  geom_qq(distribution = qnorm) +
  geom_qq_line(distribution = qnorm, col = "blue")
```

Re-expressing the data seems to have improved the *normality* of the residuals somewhat. Though not perfect, more points seem to follow a straight line. But, recall from chapter 18.4 that noise can be quite prominent in a normal q-q plot when working with small datasets.

[^1]: Cleveland, William. *Visualizing data*. Hobart Press. 1993. [pp 87-88]
