---
output: html_document
editor_options: 
  chunk_output_type: console
---
# What else can we learn from the residuals?

```{r echo=FALSE}
source("libs/Common.R")
```

```{r echo = FALSE}
pkg_ver(c("ggplot2"))
```

-----------------------------------

So far, we’ve focused on modeling the typical value of $y$ as a function of $x$. The fitted model represents a measure of location (e.g., the mean) of $y$ for infinitesimally thin slices of $x$. In the previous chapter, we used residual-dependence plots and residual-fit plots to refine the model’s fit and evaluate its accuracy.

In the univariate analysis portion of this course, we emphasized the importance of maintaining a consistent spread of residuals across groups. A uniform residual spread simplified comparisons between groups by reducing the analysis to a comparison of their means.

Similarly, ensuring a *consistent* spread of residuals across the full range of the independent variable in bivariate analysis is crucial. This consistency not only offers explanatory clarity but is also critical for many statistical procedures that assume *homoscedasticity* (constant variance) in the residuals. Violations of this assumption can compromise the validity of these methods, emphasizing the importance of carefully evaluating residual behavior during model assessment.

## The spread-location plot

While inconsistency in spread across the full range of dependent variables can be sometimes observed in a residual-dependence plot, certain patterns in the data can make such an assessment more challenging in such a plot.

A **spread-location plot** (S-L plot) is designed to explore changes in spreads as a function of increasing $x$ values. The plot pits an expression of spread, typically the square root of the residuals' absolute value, as a function of the fitted values. To help gauge the shape of this distribution, a non-parametric curve, such as the loess, is fitted to the data.

An example of a *homoscedastic* set of residuals follows. The plot on the left is the regression model and the plot on the right is the resulting residuals S-L plot.

```{r fig.width = 7, fig.height = 3, echo = FALSE}
library(tukeyedar)
set.seed(787)
n <- 100
x <- runif(n,1,10)
y <- x*2 +rnorm(n,0,0.2)
df <- data.frame(x,y)
OP<-par(mfrow=c(1,2))
 M <- eda_lm(df, x, y, show.par = FALSE, sd=FALSE, mean = FALSE)
 eda_sl(M)
 #eda_sl(M, ylab = expression(sqrt(abs(residuals))))
par(OP)
```

Here, the residuals are constant across the full range of fitted values. This is confirmed by the loess fit which shows no significant deviation from a horizontal line.

This next example is that of a model that generates a *heteroscedastic* set of residuals.

```{r fig.width = 7, fig.height = 3, echo = FALSE}
set.seed(787)
y <- x*2 +rnorm(n,0,abs(x)/3)
df <- data.frame(x,y)
OP<-par(mfrow=c(1,2))
 M <- eda_lm(df, x, y, show.par = FALSE, sd=FALSE, mean = FALSE)
 eda_sl(M, ylab = expression(sqrt(abs(residuals))))
par(OP)
```

The increasing spread as a function of increasing fitted value is apparent in the S-L plot (right-plot). It can also be observed in the $Y$ vs. $X$ plot (left plot). Note that the residual is the distance between the fitted line and each point when measured *parallel* to the $Y$ axis.

## Variation of the S-L plot

For bivariate models, an alternative to the S-L plot is the spread-dependence (S-D) plot where the independent variable, $X$, is plotted in the x-axis instead of the fitted values. This alternate form of the S-L plot is better suited for models that take on a quadratic form. For example, the following fitted model shows a monotonic increase in spread with increasing x-value. However, the S-L plot does a poor job in picking the heteroscedasticity in the residuals.

```{r fig.width = 7, fig.height = 3, echo = FALSE}
set.seed(787)
n <- 500
x <- runif(n,-5,5) 
y <- 11 + 0.2*x + 0.4 * x^2 + rnorm(n,0,abs(x+5)/5)
df <- data.frame(X=x,Y=y)
OP<-par(mfrow=c(1,2));M <- eda_lm(df, X, Y, poly=2);eda_sl(M);par(OP)
```

The heteroscedasticity in the residuals is far more pronounced when plotting  the spread as a function of the independent variable.

```{r fig.width = 7, fig.height = 3, echo = FALSE}
OP<-par(mfrow=c(1,2));M <- eda_lm(df, X, Y, poly=2)
eda_sl(M, type = "dependence");par(OP)
```

## Generating an S-L plot with `eda_sl`

If a regression model was generated using the base `lm` function or `tukeyedar`'s `eda_lm` function, the resulting model can be passed to the `eda_sl` function as follows:

```{r fig.width = 3, fig.height = 3}
library(tukeyedar)

M <- lm(mpg ~ hp, mtcars)
eda_sl(M)
```

To generate an S-D plot, set the argument `type` to `"dependence"`.

```{r fig.width = 3, fig.height = 3}
library(tukeyedar)

M <- lm(mpg ~ hp, mtcars)
eda_sl(M, type = "dependence")
```


## Generating an S-L plot with base plot or `ggplot`

Before generating an S-L plot using the base plotting environment or `ggplot`, the spread will need to be computed from the model output.


```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
library(ggplot2)

sl2 <- data.frame( std.res = sqrt(abs(residuals(M))), 
                   fit     = predict(M))

ggplot(sl2, aes(x = fit, y  =std.res)) + geom_point() +
              stat_smooth(method = "loess", se = FALSE, span = 1, 
                          method.args = list(degree = 1) ) +
              ylab(expression(sqrt(abs(residuals)))) +
              xlab("Fitted values")
```

The function `predict()` extracts the fitted y-values from the model `M` and is plotted along the x-axis. 


We learned with the univariate data that re-expressing values was one way to correct for any increasing or decreasing spreads across the range of fitted values. At this point, we may want to look into re-expressing the data. But, before we do, we'll explore another diagnostic plot: the *normal q-q plot*.

### Checking residuals for normality

If you are interested in conducting a hypothesis test (i.e. addressing the question *"is the slope significantly different from 0"*) you will likely want to check the residuals for normality since this is an assumption made when computing a confidence interval and a p-value. 

In `ggplot`, we learned that a normal q-q plot could be generated using the `stat_qq` and `stat_qq_line` function.

```{r fig.height=2.5, fig.width=2.5}
ggplot(df, aes(sample = residuals2)) + 
  stat_qq(distribution = qnorm) +
  stat_qq_line(distribution = qnorm, col = "blue")
```

Here, the residuals seem to stray a little from a normal distribution.

## Re-expressing the data

The monotone spread can be problematic if we are to characterize the spread of `cp.ratio` as being the same (homogeneous) across the full range of `area` values. To remedy this, we can re-express the `cp.ratio` values. Variables measured as ratios (such is the case with `cp.ratio`) are good candidates for log transformation. We will therefore fit a new linear model to the data after transforming the y-value. 

```{r}
df.log <- data.frame( area = df$area, cp.ratio.log = log(df$cp.ratio))
M3     <- lm(cp.ratio.log ~ area, dat = df.log)
```

Transforming a variable, whether $x$ or $y$, will change the nature of the relationship between both variables. We therefore need to recreate the scatter plot. We'll also see how well a 1^st^ order polynomial can characterize this relationship (re-expressing values can sometimes help "straighten" a relationship between variables, hence why we are not starting off first with a 2^nd^ order polynomial fit).

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
ggplot(df.log, aes(x = area, y = cp.ratio.log)) + geom_point() +
  stat_smooth(method = "lm", se = FALSE)
```
 
At first glance, the log transformation seems to have done a good job at straightening the batch of values. Next, let's look at the residual-dependence plot.

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
df.log$residuals <- residuals(M3)
ggplot(df.log, aes(x = area, y = residuals)) + geom_point() +
               stat_smooth(method = "loess", se = FALSE, span = 1, 
                           method.args = list(degree = 1) )
```

Logging the y values has eliminated the residual's dependence on `area`. Next, let's assess homogeneity in the residuals using the s-l plot.

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
sl3 <- data.frame( std.res = sqrt(abs(residuals(M3))), 
                   fit     = predict(M3))
ggplot(sl3, aes(x = fit, y = std.res)) + geom_point() +
               stat_smooth(method ="loess", se = FALSE, span = 1, 
                           method.args=list(degree = 1) )
```

Even though we observe a mid range peak in spread, we do not observe a **systematic increase** in spread. The log transformation seems to have removed the monotone spread. 

Finally, we'll check for normality of the residuals.

```{r fig.height=2.5, fig.width=2.5}
ggplot(df.log, aes(sample = residuals)) + 
  geom_qq(distribution = qnorm) +
  geom_qq_line(distribution = qnorm, col = "blue")
```

Re-expressing the data seems to have improved the *normality* of the residuals somewhat. Though not perfect, more points seem to follow a straight line. But, recall from chapter 18.4 that noise can be quite prominent in a normal q-q plot when working with small datasets.

[^1]: Cleveland, William. *Visualizing data*. Hobart Press. 1993. [pp 87-88]
