---
output: html_document
editor_options: 
  chunk_output_type: console
---
# What else can we learn from the residuals?

```{r echo=FALSE}
source("libs/Common.R")
```

```{r echo = FALSE}
pkg_ver(c("ggplot2"))
```

-----------------------------------

So far, we’ve focused on modeling the typical value of $y$ as a function of $x$. The fitted model represents a measure of location (e.g., the mean) of $y$ for infinitesimally thin slices of $x$. In the previous chapter, we used residual-dependence plots and residual-fit plots to refine the model’s fit and evaluate its accuracy.

In the univariate analysis portion of this course, we emphasized the importance of maintaining a consistent spread of residuals across groups. A uniform residual spread simplified comparisons between groups by reducing the analysis to a comparison of their means.

Similarly, ensuring a *consistent* spread of residuals across the full range of the independent variable in bivariate analysis is crucial. This consistency not only offers explanatory clarity but is also critical for many statistical procedures that assume *homoscedasticity* (constant variance) in the residuals. Violations of this assumption can compromise the validity of these methods, emphasizing the importance of carefully evaluating residual behavior during model assessment.

## The spread-location plot

While inconstancy in spread across the full range of dependent variables can be sometimes observed in a residual-dependence plot, certain patterns in the data can make such an assessment more challenging in such a plot.

A **spread-location plot** (SL plot) is designed to explore patterns in changes in spreads as a function of increasing $x$ values. The plot pits an expression of spread, typically the square root of the residuals' absolute value as a function of the fitted values. To help gauge the shape of this distribution, a non-parametric curve, such as the loess, is fitted to the data.

An example of a *homoscedastic* set of residuals follows. The plot on the left is the regression model and the plot on the right is the resulting residuals SL plot.

```{r fig.width = 7, fig.height = 3, echo = FALSE}
set.seed(787)
n <- 100
x <- runif(n,1,10)
y <- x*2 +rnorm(n,0,0.2)
df <- data.frame(x,y)
OP<-par(mfrow=c(1,2))
 M <- eda_lm(df, x, y, show.par = FALSE, sd=FALSE, mean = FALSE)
 eda_sl(M, ylab = expression(sqrt(abs(residuals))))
par(OP)
```

Here, the residuals are constant across the full range of fitted values. This is confirmed by the loess fit which shows no significant deviation from a horizintal line.

In this next example, an example of a model that generates a heteroscedastic set of residuals is shown.

```{r fig.width = 7, fig.height = 3, echo = FALSE}
set.seed(787)
y <- x*2 +rnorm(n,0,abs(x)/3)
df <- data.frame(x,y)
OP<-par(mfrow=c(1,2))
 M <- eda_lm(df, x, y, show.par = FALSE, sd=FALSE, mean = FALSE)
 eda_sl(M, ylab = expression(sqrt(abs(residuals))))
par(OP)
```

The increasing spread as a function of increasing fitted value is apparent in the SL plot (right-plot). It can also be observed in the $Y$ vs. $X$ plot (left plot). Note that the residual is the distance between the fitted line and each point when measured *parallel* to the $Y$ axis.

## Generating 

The `M2` and `lo` models do a good job in eliminating any dependence between residual and x-value. Next, we will check that **the residuals do not show a dependence with *fitted* y-values**. This is analogous to univariate analysis where we checked if residuals increased or decreased with increasing medians across categories. Here we will compare residuals to the fitted `cp.ratio` values (for a univariate analogy, think of the fitted line as representing a *level* across different segments along the x-axis). We'll generate a spread-level plot of model `M2`'s residuals (note that in the realm of regression analysis, such plot is often referred to as a **scale-location** plot). We'll also add a loess curve to help visualize any patterns in the plot.

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
sl2 <- data.frame( std.res = sqrt(abs(residuals(M2))), 
                   fit     = predict(M2))

ggplot(sl2, aes(x = fit, y  =std.res)) + geom_point() +
              stat_smooth(method = "loess", se = FALSE, span = 1, 
                          method.args = list(degree = 1) )
```

The function `predict()` extracts the y-values from the fitted model `M2` and is plotted along the x-axis. It's clear from this plot that the residuals are not homogeneous; they increase as a function of increasing *fitted* CP ratio. The "bend" observed in the loess curve is most likely due to a single point at the far (right) end of the fitted range. Given that we have a small batch of numbers, a loess can be easily influenced by an outlier. We may therefore want to increase the loess span by setting `span = 2`.


```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
ggplot(sl2, aes(x = fit, y = std.res)) + geom_point() +
              stat_smooth(method = "loess", se = FALSE, span = 2, 
                          method.args = list(degree = 1) )
```

The point's influence is reduced enough to convince us that the observed monotonic increase is real. 

We learned with the univariate data that re-expressing values was one way to correct for any increasing or decreasing spreads across the range of fitted values. At this point, we may want to look into re-expressing the data. But, before we do, we'll explore another diagnostic plot: the *normal q-q plot*.

### Checking residuals for normality

If you are interested in conducting a hypothesis test (i.e. addressing the question *"is the slope significantly different from 0"*) you will likely want to check the residuals for normality since this is an assumption made when computing a confidence interval and a p-value. 

In `ggplot`, we learned that a normal q-q plot could be generated using the `stat_qq` and `stat_qq_line` function.

```{r fig.height=2.5, fig.width=2.5}
ggplot(df, aes(sample = residuals2)) + 
  stat_qq(distribution = qnorm) +
  stat_qq_line(distribution = qnorm, col = "blue")
```

Here, the residuals seem to stray a little from a normal distribution.

## Re-expressing the data

The monotone spread can be problematic if we are to characterize the spread of `cp.ratio` as being the same (homogeneous) across the full range of `area` values. To remedy this, we can re-express the `cp.ratio` values. Variables measured as ratios (such is the case with `cp.ratio`) are good candidates for log transformation. We will therefore fit a new linear model to the data after transforming the y-value. 

```{r}
df.log <- data.frame( area = df$area, cp.ratio.log = log(df$cp.ratio))
M3     <- lm(cp.ratio.log ~ area, dat = df.log)
```

Transforming a variable, whether $x$ or $y$, will change the nature of the relationship between both variables. We therefore need to recreate the scatter plot. We'll also see how well a 1^st^ order polynomial can characterize this relationship (re-expressing values can sometimes help "straighten" a relationship between variables, hence why we are not starting off first with a 2^nd^ order polynomial fit).

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
ggplot(df.log, aes(x = area, y = cp.ratio.log)) + geom_point() +
  stat_smooth(method = "lm", se = FALSE)
```
 
At first glance, the log transformation seems to have done a good job at straightening the batch of values. Next, let's look at the residual-dependence plot.

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
df.log$residuals <- residuals(M3)
ggplot(df.log, aes(x = area, y = residuals)) + geom_point() +
               stat_smooth(method = "loess", se = FALSE, span = 1, 
                           method.args = list(degree = 1) )
```

Logging the y values has eliminated the residual's dependence on `area`. Next, let's assess homogeneity in the residuals using the s-l plot.

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
sl3 <- data.frame( std.res = sqrt(abs(residuals(M3))), 
                   fit     = predict(M3))
ggplot(sl3, aes(x = fit, y = std.res)) + geom_point() +
               stat_smooth(method ="loess", se = FALSE, span = 1, 
                           method.args=list(degree = 1) )
```

Even though we observe a mid range peak in spread, we do not observe a **systematic increase** in spread. The log transformation seems to have removed the monotone spread. 

Finally, we'll check for normality of the residuals.

```{r fig.height=2.5, fig.width=2.5}
ggplot(df.log, aes(sample = residuals)) + 
  geom_qq(distribution = qnorm) +
  geom_qq_line(distribution = qnorm, col = "blue")
```

Re-expressing the data seems to have improved the *normality* of the residuals somewhat. Though not perfect, more points seem to follow a straight line. But, recall from chapter 18.4 that noise can be quite prominent in a normal q-q plot when working with small datasets.

[^1]: Cleveland, William. *Visualizing data*. Hobart Press. 1993. [pp 87-88]
