---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Spread-location plot

```{r echo=FALSE}
source("libs/Common.R")
```

```{r echo = FALSE}
pkg_ver(c("ggplot2"))
```

-----------------------------------

So far, we've learned how the typical value of $y$ changes as a function of $x$. The fitted model can be thought of as a measure of location (e.g. the mean) of $y$ for different infinitesimally thin *slices* of $x$. We made use of the residual dependence plot to help fine tune the model.

In the univariate analysis portion of the course, we learned of the desirability of maintaining a consistent spread in residuals across groups. Doing so reduced the comparison of groups to a comparison of means. 

The desirability in maintaining a constiant spread in residuals across the full range of teh dependent variable is equally important given that many statitical procedures rely on this assumption.


## The spread-location plot




A bivariate model is deemed "well defined" when its residuals are constant over the full range of $x$. In essence, we expect a formula  of the form:

$$
Y = a + bx + \epsilon
$$

where $\epsilon$ is a constant that does not vary as a function of varying $x$.  This should sound quite familiar given that we've spent a good part of the univariate analysis section seeking a homogeneous spread in the residuals (i.e. a spread that did not change as a function of fitted group values). So what other diagnostic plots can we (and *should we*) generate from the fitted model? We explore other diagnostic plots next.


### Spread-location plot

The `M2` and `lo` models do a good job in eliminating any dependence between residual and x-value. Next, we will check that **the residuals do not show a dependence with *fitted* y-values**. This is analogous to univariate analysis where we checked if residuals increased or decreased with increasing medians across categories. Here we will compare residuals to the fitted `cp.ratio` values (for a univariate analogy, think of the fitted line as representing a *level* across different segments along the x-axis). We'll generate a spread-level plot of model `M2`'s residuals (note that in the realm of regression analysis, such plot is often referred to as a **scale-location** plot). We'll also add a loess curve to help visualize any patterns in the plot.

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
sl2 <- data.frame( std.res = sqrt(abs(residuals(M2))), 
                   fit     = predict(M2))

ggplot(sl2, aes(x = fit, y  =std.res)) + geom_point() +
              stat_smooth(method = "loess", se = FALSE, span = 1, 
                          method.args = list(degree = 1) )
```

The function `predict()` extracts the y-values from the fitted model `M2` and is plotted along the x-axis. It's clear from this plot that the residuals are not homogeneous; they increase as a function of increasing *fitted* CP ratio. The "bend" observed in the loess curve is most likely due to a single point at the far (right) end of the fitted range. Given that we have a small batch of numbers, a loess can be easily influenced by an outlier. We may therefore want to increase the loess span by setting `span = 2`.


```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
ggplot(sl2, aes(x = fit, y = std.res)) + geom_point() +
              stat_smooth(method = "loess", se = FALSE, span = 2, 
                          method.args = list(degree = 1) )
```

The point's influence is reduced enough to convince us that the observed monotonic increase is real. 

We learned with the univariate data that re-expressing values was one way to correct for any increasing or decreasing spreads across the range of fitted values. At this point, we may want to look into re-expressing the data. But, before we do, we'll explore another diagnostic plot: the *normal q-q plot*.

### Checking residuals for normality

If you are interested in conducting a hypothesis test (i.e. addressing the question *"is the slope significantly different from 0"*) you will likely want to check the residuals for normality since this is an assumption made when computing a confidence interval and a p-value. 

In `ggplot`, we learned that a normal q-q plot could be generated using the `stat_qq` and `stat_qq_line` function.

```{r fig.height=2.5, fig.width=2.5}
ggplot(df, aes(sample = residuals2)) + 
  stat_qq(distribution = qnorm) +
  stat_qq_line(distribution = qnorm, col = "blue")
```

Here, the residuals seem to stray a little from a normal distribution.

## Re-expressing the data

The monotone spread can be problematic if we are to characterize the spread of `cp.ratio` as being the same (homogeneous) across the full range of `area` values. To remedy this, we can re-express the `cp.ratio` values. Variables measured as ratios (such is the case with `cp.ratio`) are good candidates for log transformation. We will therefore fit a new linear model to the data after transforming the y-value. 

```{r}
df.log <- data.frame( area = df$area, cp.ratio.log = log(df$cp.ratio))
M3     <- lm(cp.ratio.log ~ area, dat = df.log)
```

Transforming a variable, whether $x$ or $y$, will change the nature of the relationship between both variables. We therefore need to recreate the scatter plot. We'll also see how well a 1^st^ order polynomial can characterize this relationship (re-expressing values can sometimes help "straighten" a relationship between variables, hence why we are not starting off first with a 2^nd^ order polynomial fit).

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
ggplot(df.log, aes(x = area, y = cp.ratio.log)) + geom_point() +
  stat_smooth(method = "lm", se = FALSE)
```
 
At first glance, the log transformation seems to have done a good job at straightening the batch of values. Next, let's look at the residual-dependence plot.

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
df.log$residuals <- residuals(M3)
ggplot(df.log, aes(x = area, y = residuals)) + geom_point() +
               stat_smooth(method = "loess", se = FALSE, span = 1, 
                           method.args = list(degree = 1) )
```

Logging the y values has eliminated the residual's dependence on `area`. Next, let's assess homogeneity in the residuals using the s-l plot.

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
sl3 <- data.frame( std.res = sqrt(abs(residuals(M3))), 
                   fit     = predict(M3))
ggplot(sl3, aes(x = fit, y = std.res)) + geom_point() +
               stat_smooth(method ="loess", se = FALSE, span = 1, 
                           method.args=list(degree = 1) )
```

Even though we observe a mid range peak in spread, we do not observe a **systematic increase** in spread. The log transformation seems to have removed the monotone spread. 

Finally, we'll check for normality of the residuals.

```{r fig.height=2.5, fig.width=2.5}
ggplot(df.log, aes(sample = residuals)) + 
  geom_qq(distribution = qnorm) +
  geom_qq_line(distribution = qnorm, col = "blue")
```

Re-expressing the data seems to have improved the *normality* of the residuals somewhat. Though not perfect, more points seem to follow a straight line. But, recall from chapter 18.4 that noise can be quite prominent in a normal q-q plot when working with small datasets.

[^1]: Cleveland, William. *Visualizing data*. Hobart Press. 1993. [pp 87-88]
