---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Bivariate residual-fit spread plot

```{r echo=FALSE}
source("libs/Common.R")
```

```{r echo = FALSE}
pkg_ver(c("ggplot2", "tukeyedar"))
```

-----------------------------------

William Cleveland's *residual-fit spread* (rfs) plot was introduced earlier in this course in the univariate context. The plot provides a visual means of comparing the variability explained by a fitted model—such as group means—to the variability of the residuals. The rfs plot can also be applied to bivariate models, offering insight into the relationship between a model's effects and its residual variability.

In bivariate analysis, an important objective is to identify a model that maximizes our ability to explain the variability in $Y$. The greater the variability in $Y$ explained by the model, the smaller the uncertainty in estimating $Y$ for each value of $X$. This objective translates into minimizing the spread of residuals relative to the variability captured by the model.

For example, in the figure below, the plot on the left illustrates a model that effectively explains the variability in $Y$. Notice the relatively small residuals, represented as vertical dashed lines connecting the fitted (predicted) values (red points) to the observed values (grey points), compared to the variability explained by the model. The model accounts for a range in 
$Y$ of approximately 11 units, while the residual magnitudes average about 1 unit, indicating minimal unexplained variability.

In contrast, the plot on the right shows a model with limited explanatory power. Here, the model accounts for only about 2 units of variability in $Y$, while the residual magnitudes average approximately 4 units. This suggests that a significant portion of the variability in $Y$ remains unexplained by $X$.


```{r echo = FALSE, results='hide', fig.width=8, fig.height=3}
library(tukeyedar)

n <- 20
set.seed(123)
x <- sort(runif(n, 1, 10))
y <- 1 + 1.5 * x + rnorm(n)
#y[20] <- 28
df <- data.frame(x,y)


OP <- par(mfrow = c(1,2))
M1 <- eda_lm(df, x, y, show.par = FALSE, mean.l = FALSE, sd = FALSE)
error_segs <- lapply(1:nrow(df),
       FUN = \(i) data.frame(x = df$x[i], y = c(M1$fitted.values[i], df$y[i]),
                             row.names = NULL))
OP2 <- par(mar = c(0,0,0,0))
  lapply(error_segs, \(x) lines(x, col = "grey", lty = 2))
  points(x = df$x, y = M1$fitted.values, pch = 16, col = "darkred")
par(OP2)

set.seed(17)
y<- 10 + runif(n, -6, 6)
df <- data.frame(x,y)

M2 <-eda_lm(df,x,y, robust = TRUE, show.par = FALSE, mean.l = FALSE, sd = FALSE)
error_segs <- lapply(1:nrow(df),
       FUN = \(i) data.frame(x = df$x[i], y = c(M2$fitted.values[i], df$y[i]),
                             row.names = NULL))
OP2 <- par(mar = c(0,0,0,0))
  lapply(error_segs, \(x) lines(x, col = "grey", lty = 2))
  points(x = df$x, y = M2$fitted.values, pch = 16, col = "darkred")
par(OP2)

par(OP)
```

While the assessment of the model's power in explaining the variability can be gleaned from a scatter plot given extreme scenarios as those shown above, such a plot is not ideally suited in making this assesment in a wide range of scenarios. An RFS plot is better suited for making such an assesmnet

The next section 

## Constructing an residual-fit spread plot

For instance, consider estimating the miles-per-gallon (`mpg`) of a vehicle using the `mtcars` dataset in. A straightforward starting point is the simplest estimate: the mean `mpg`.


```{r echo=FALSE, fig.height=3, fig.width=2.5}
library(tukeyedar)

eda_jitter(mtcars, mpg)

```

There is, as expected, some variability that covers a range of about 24 miles-per-gallon.

We can refine this estimate by incorporating an additional variable into the analysis. Instead of introducing a categorical variable, as demonstrated earlier in this tutorial, we can include another continuous variable to enhance the estimate. For example, we might examine whether a vehicle's horsepower provides further insight into predicting its miles-per-gallon (mpg). The following plot is a scatter plot of `mpg` against `hp` and the fitted 1^st^ order model.

```{r echo=FALSE, fig.height=3, fig.width=2.5, results='hide'}
eda_lm(mtcars, hp, mpg, show.par = FALSE, mean.l = FALSE, sd = FALSE)
```




