---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Bivariate residual-fit spread plot

```{r echo=FALSE}
source("libs/Common.R")
```

```{r echo = FALSE}
pkg_ver(c("ggplot2", "tukeyedar"))
```

-----------------------------------

William Cleveland's *residual-fit spread* (RFS) plot was introduced earlier in this course in the univariate context. The plot provides a visual means of comparing the variability explained by a fitted model—such as group means—to the variability of the residuals. the RFS plot can also be applied to bivariate models, offering insight into the relationship between a model's effects and its residual variability.

As with univariate analysis, the goal is to estimate a variable while minimizing the uncertainty associated with that estimate. For instance, consider estimating the miles-per-gallon (`mpg`) of a vehicle using the `mtcars` dataset in. A straightforward starting point is the simplest estimate: the mean `mpg`.

```{r echo=FALSE, fig.height=3, fig.width=2.5}
library(tukeyedar)

eda_jitter(mtcars, mpg)

```

There is, as expected, some variability that covers a range of about 24 miles-per-gallon.

We can refine this estimate by incorporating an additional variable into the analysis. Instead of introducing a categorical variable, as demonstrated earlier in this tutorial, we can include another continuous variable to enhance the estimate. For example, we might examine whether a vehicle's horsepower provides further insight into predicting its miles-per-gallon (mpg). The following plot is a scatter plot of `mpg` against `hp` and the fitted 1^st^ order model.

```{r echo=FALSE, fig.height=3, fig.width=2.5, results='hide'}
eda_lm(mtcars, hp, mpg, show.par = FALSE, mean.l = FALSE, sd = FALSE)
```




