# Preface {.unnumbered}


This book is a compilation of lecture notes used in an *Exploratory Data Analysis in R* course taught to undergraduates at Colby College. The course assumes little to no background in quantitative analysis nor in computer programming and was first taught in Spring, 2015. The course introduces students to data manipulation in R, data exploration (in the spirit of John Tukey's EDA) and the R markdown language. Many of the visualization techniques are adopted from William Cleveland's *Data Visualization* book.

The base R plotting environment and the `ggplot2` ecosystem are used throughout this book. While a chapter is dedicated to the `lattice` plotting package, its functions are not used outside of that chapter given that `ggplot2` offers many of `lattice`'s functionality.

While great effort is made to adopt a consist plotting environment throughout this book (this being `ggplot2`, for the most part), a few topics (including the q-q plot and the median polish) will benefit from custom plotting functions available in the `tukeyedar` package. The [package](https://github.com/mgimond/tukeyedar/) can be downloaded from GitHub via the command:

```{r eval = FALSE}
devtools::install_github("mgimond/tukeyedar")
```

Note that installing the GitHub package will require that the `devtools` package be installed first.

Functions making use of the `tukeyedar` package will be highlighted in a peach/pink code block as opposed to the default light yellow code block used for all other code blocks. For example, if `tukeyedar`'s `eda_qq` function is used, the code block will take on the following appearance:

```{r class.source="eda", eval=FALSE}
library(tukeyedar)
eda_qq(Tenor, Bass)
```

The `tukeyedar` functions are built off of base R graphics and require R vesion `4.1` or greater.

------------------

Manuel "Manny" Gimond





