---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Univariate spread-level plot


```{r echo=FALSE}
source("libs/Common.R")
```


```{r echo = FALSE}
pkg_ver(c("dplyr", "ggplot2", "tukeyedar"))
```

-----

In Chapter 19, we emphasized the importance of maintaining a homogeneous spread in residuals. One common cause of non-homogeneous spread is when the dataset exhibits a systematic change in variability as a function of location. In other words, the variability within each batch may depend on a location-based statistic, such as the mean or median. This type of dependency is often undesirable—especially in analyses like ANOVA—and should ideally be addressed.

To visualize this relationship, a **spread-level plot** (s-l)--sometimes referred to as a spread-location plot--is highly effective. This plot helps identify patterns or trends in variability that may need correction before proceeding with further analysis.

## Constructing the (univariate) s-l plot

The s-l plot visualizes the relationship between the residuals and the location for each batch of data (typically the median, due to its robustness against outliers) . The residuals are expressed as the square root of their absolute value. 

For example, the following is an s-l plot of petal length vs species from the `iris` dataset.

```{r , fig.height=3,fig.width=4, echo = FALSE}
library(tukeyedar)
eda_sl(iris, Petal.Length, Species, show.par = FALSE)
```

The red line in the plot connects the median values of each batch of residuals. It helps identify the type of relationship between spread and location. If the line increases *monotonically upward*, there is an increasing spread as a function of increasing location; if the line decreases *monotonically downward*, there is a decreasing spread as a function of increasing location; and if the line is neither increasing nor decreasing monotonically, there is no change in spread as a function of location.

The s-l plot in the `iris` dataset suggests that the residuals increase monotonically with an increase in the fitted median value. Note that the x-axis is not categorical, it shows the median values for each species group. 

Though not as effective in highlighting  the heterogeneity in spread as in the s-l plot, the increase in spread can be observed in a boxplot of the data.

```{r, fig.height=3,fig.width=4, echo = FALSE}
eda_boxls(iris, Petal.Length, Species, show.par = FALSE)
```

Here, the increasing width of the IQRs is also noticeable.

Increasing or decreasing spreads as a function of fitted value can be corrected by re-expressing the continuous variable. For example, applying a log transformation helps stabilize the residuals as can be seen in the following s-l plot. 

```{r, fig.height=3,fig.width=4, echo = FALSE}
eda_sl(iris, Petal.Length, Species, p = 0, show.par = FALSE)
title(expression("log(petal length)"), col.main = "grey")
```

Note that the re-expression of the petal length values changes the group median values but not the rank of species.

## Creating an s-l plot using `eda_sl`

An s-l plot can be generated using `tukeyedar`'s `eda_sl` function.

```{r class.source="eda" , fig.height=3,fig.width=4}
eda_sl(iris, Petal.Length, Species)
```

As with many `tukeyedar` functions used in the course, the power transformation applied to the data is shown in the upper right-hand corner of the plot. By default, the power transformation is `1` (i.e. an untransformed data).

The function allows you to apply the transformation without needing to do so outside of the function. By default, the Box-Cox transformation method is adopted. To adopt the Tukey method, set `Tukey = TRUE`.

In this following code block, we apply a power transformation of `0` (the log transformation) to the data.

```{r class.source="eda" , fig.height=3,fig.width=4}
eda_sl(iris, Petal.Length, Species, p = 0)
```

## Creating an s-l plot with `ggplot`

The following code block demonstrates the steps required to create an s-l plot in `ggplot`.

```{r, fig.height=2.5,fig.width=4}
library(dplyr)
library(ggplot2)

res.sq <-  iris %>% 
  group_by(Species) %>% 
  mutate(Median   = median(Petal.Length),
         Residual = sqrt(abs(Petal.Length - Median)))

ggplot(res.sq, aes(x=Median, y=Residual)) + 
  geom_jitter(alpha=0.4, width=0.05, height=0) +
  stat_summary(fun = median, geom = "line", col = "red") +
  ylab("Spread") +
  geom_text(aes(x = Median, y = 1.25, label = Species)) +
  xlim(1, 6.5)
```

> Note that if you are to rescale the y-axis when using the `stat_summary()` function, you should use the `coord_cartesian(ylim = c( .. , .. ))` function instead of the `ylim()` function. The latter will mask the values above its maximum range from the `stat_summary()` function, the former will not.

## A variation of the s-l plot

Another version of the s-l plot pits the *log of the inter-quartile spread* against the *log of the median*. This approach only works for positive values (this may require that values be adjusted so that the minimum value be greater than 0).  

This variation of the s-l plot is appealing in that the slope of the best fit line can suggest a power transformation via $power = 1 - slope$.

This variant of the s-l plot can be implemented in the `eda_sl` function by setting the argument `type = "level"`.

```{r class.source="eda" , fig.height=3,fig.width=3}
eda_sl(iris, Petal.Length, Species, type = "level",
       loess.d = list(degree = 1, span = 1.5))
```

Note how this plot differs from our earlier s-l plot in that we are only displaying each batch's median spread value and we are *fitting* a straight line to the medians instead of *connecting* them.

The function will return the slope of the line in the console window. Here, the slope is 1.05 which suggests a power of $1 - 1.05 = -.05$. This is a power transformation very close to the log transformation used earlier in this chapter.

A `ggplot` implementation of this variant of the s-l plot is shown next:

```{r fig.height=2.8, fig.width=2.8}
sl <- iris %>%
  group_by(Species)  %>%
  summarise (level  = log(median(Petal.Length)),
                IQR = IQR(Petal.Length),  # Computes the interquartile range
             spread = log(IQR))

ggplot(sl, aes(x = level, y = spread)) + geom_point() + 
  stat_smooth(method = MASS::rlm, se = FALSE) +
  xlab("Location") + ylab("Spread") +
  geom_text(aes(x = level, y = spread, label = Species), cex=2.5)
```


