---
output: html_document
editor_options: 
  chunk_output_type: console
---
# The theoretical q-q plot


```{r echo=FALSE}
source("libs/Common.R")
```


```{r echo = FALSE}
pkg_ver(c("dplyr", "ggplot2","lattice", "tukeyedar"))
```


## Introduction

Thus far, we have used the quantile-quantile plots to compare the distributions between two empirical (i.e. observational) datasets--hence the name *empirical q-q plot*. We can also use the q-q plot to compare an empirical distribution to a theoretical distribution (i.e. one defined mathematically). Such a plot is usually referred to as a *theoretical Q-Q plot*. Examples of popular theoretical distribution are the normal distribution (aka the Gaussian distribution), the chi-square distribution, and the exponential distribution just to name a few.

```{r, fig.height=2, fig.width=7, echo=FALSE}
OP <- par( mfrow=c(1,4))
x <- seq(-5, 5, 0.1)
plot(dnorm(x,0,1) ~ x, type = "l",axes = FALSE, ylab = NA, xlab = "Normal")
x <- seq(0, 20, 0.1)
plot(dchisq(x,3,1) ~ x, type = "l",axes = FALSE, ylab = NA, xlab = "Chi-square")
x <- seq(0, 10, 0.1)
plot(dexp(x,1) ~ x, type = "l",axes = FALSE, ylab = NA, xlab = ("Exponential"))
par(OP)
```



There are many reasons we might want to compare empirical data to theoretical distributions:

 * A theoretical distribution is easy to parameterize. For example, if the shape of the distribution of a batch of numbers can be approximated by a normal distribution we can reduce the complexity of our data to just two values: the mean and the standard deviation.
 
 * If data can be approximated by certain theoretical distributions, then many mainstream statistical procedures can be applied to the data.
 
 * In inferential statistics, knowing that a sample was derived from a population whose distribution follows a theoretical distribution allows us to derive certain properties of the population from the sample. For example, if we know that a sample comes from a normally distributed population, we can define confidence intervals for the sample mean using a t-distribution.
 
 * Modeling the distribution of the observed data can provide insight into the underlying process that generated the data.
 
But very few empirical datasets follow any theoretical distributions exactly. So the questions usually ends up being "how well does theoretical distribution *X* fit my data?"

The theoretical quantile-quantile plot is a tool to explore how a batch of numbers deviates from a theoretical distribution and to visually assess whether the difference is significant for the purpose of the analysis. In the following examples, we will compare empirical data to the normal distribution using the **normal quantile-quantile** plot or **normal q-q** plot for short.

## The normal q-q plot

The *normal q-q plot* is just a special case of the *empirical q-q plot* we've covered in the previous chapter, the difference being that we assign the empirical quantiles to the y-axis and the matching normal quantiles to the x-axis. 


```{r echo=FALSE, fig.height=3, fig.width=3}
library(tukeyedar)
set.seed(121)
x <- rnorm(30) * 2 + 32
eda_qq(x, norm=T, show.par = FALSE, q = FALSE)
```

When comparing a batch of numbers to a theoretical distribution on a q-q plot, we are looking for significant deviation from a straight line. To make it easier to judge straightness, we can fit a line to the points. Note that we are **not** creating a 45&deg; (x=y) slope as was done with the empirical q-q plot--the range of values between both sets of numbers do not match. Here, we are only seeking the **straightness** of the point pattern.


We'll first learn how to generate this plot using the built-in R function, then we'll do the same with `ggplot2`.


### Using R’s built-in functions

In the following example, we'll compare the `Alto 1` group to a normal distribution. 

```{r}
library(dplyr)

df <- lattice::singer

alto  <- df %>% 
  filter(voice.part == "Alto 1") %>% 
  pull(height)
```

Note that `alto` is a single vector element. We'll use two built-in functions to generate a normal q-q plot: `qqnorm` and `qqline`. 

```{r fig.height=3, fig.width=3, echo=2:3, results='hold'}
OP <-par(mar=c(4,4,1,1))
qqnorm(alto)
qqline(alto, qtype = 5)
par(OP)
```

There are many ways one can fit a line to the data, Here, we opt to fit a line to the first and third quartiles (IQR) of the q-q plot. We are also adopting the [f-value type of `5`](univariate_plots.html#computing-the-f-quantile) by setting `qtype = 5` in the call to `qqline`. 


### Using the ggplot2 plotting environment

To generate the theoretical q-q plot, we first use the `stat_qq` function to generate the point plot, then we call the `stat_qq_line` function to generate the IQR fit. Here, we are passing a single vector object instead of a dataframe to `ggplot()`.

```{r, fig.height=3, fig.width=3}
library(ggplot2)

ggplot() + aes(sample = alto) + stat_qq(distribution = qnorm) +
  stat_qq_line(col = "blue") +
  xlab("Unit normal quantile") + ylab("Height")

```

Note the slight difference in syntax used with `ggplot` when passing a vector instead of a dataframe to the function. Here, we take the `aes()` function outside of the `ggplot()` function. This is done to render a cleaner syntax. The alternative, `ggplot(,aes(sample = alto))`, would make it difficult to notice the comma just before `aes()`, thus increasing the chance for a typo.

The `stat_qq_line` function uses the built-in `quantile` function and, as such, will adopt the default quantile type `7` (i.e. it computes the f-value as $(i - 1)/(n - 1))$. This differs from our approach to computing the f-value. This explains the slight difference in the fitted line between the `ggplot` method and the base method we used earlier where we set `qtype` to `5`. This setting cannot be changed in `stat_qq_line`. However, the difference in methods has little to no impact in our assessment of the "straightness" of the points.

Note that `geom_qq` and  `geom_qq_line` functions are identical to `stat_qq`and `stat_qq_line`.

## How normal is my dataset?

The `alto` batch of values seem to do a good job in following a normal distribution given how well they follow a straight line. The stair-step pattern in the points is simply a byproduct of the rounding of height values to the nearest inch. A few observations at the tail ends of the distribution deviate from normal, but this is to be expected given that tail ends of distributions tend to be noisy.

So how do the other singer groups compare to a normal distribution? We'll make use of ggplot's faceting function to generate all eight normal q-q plots.

```{r fig.height=2.3, fig.width=8}
ggplot(df, aes(sample=height)) + stat_qq(distribution=qnorm) +
  stat_qq_line( col = "blue") +
  xlab("Unit normal quantile") + ylab("Height") +
  facet_wrap(~voice.part, nrow = 1)
```

For the most part, all eight batches appear to follow a normal distribution.

## What would a dataset pulled from a normal distribution look like?

Simulations are a great way to develop an intuitive feel for what a dataset pulled from a normal distribution might look like in a normal q-q plot. You will seldom come across perfectly normal data in the real world. Noise is an inherent part of any underlying process. As such, random noise can influence the shape of a q-q plot despite the data coming from a normal distribution. This is especially true with small datasets as demonstrated in the following example where we simulate five small batches of values pulled from a normal distribution. The `rnorm` function is used in this example to randomly pick a number from a normal distribution whose mean is set to the mean of the `alto` values  and whose standard deviation is set to the  standard deviation of the `alto` values. We also round the values to mimic the rounding of height values observed in the `singer` dataset.


```{r fig.height=2.5, fig.width=9}
set.seed(321)  # Sets random generator seed for consistent output

# Simulate values from a normal distribution
sim <- data.frame(sample = paste0("Sample",1:5),
                  value  = round(rnorm(length(alto)*5, 
                                 mean = mean(alto), sd = sd(alto))))

# Generate q-q plots of the simulated values
ggplot(sim, aes(sample = value)) + stat_qq(distribution = qnorm) +
  stat_qq_line(line.p = c(0.25, 0.75), col = "blue") +
  xlab("Unit normal quantile") + ylab("Simulated normals") +
  facet_wrap(~ sample, nrow = 1) 

```


Of the five simulated batches, `Sample3` generates a *textbook*  normal q-q plot that one would expect from a normally distributed batch of values. `Sample2` *could* lead one to question whether the data were pulled from a normal distribution, even though we know that they were!

The singer height normal q-q plots do not look different from some of these simulated plots. In fact, they probably look more *Normal* then the simulated set of values! This lends confidence in our earlier verdict that the `singer` height distributions can be characterized by a normal distribution.

## How normal q-q plots behave in the face of skewed data

It can be helpful to simulate distributions of difference skewness to see how a normal quantile plot may behave. In the following figure, the top row shows different density distribution plots; the bottom row shows the **normal q-q plots** for each distribution.


```{r, echo=FALSE, fig.width=9,fig.height=4}
# q.q function
# =============
# Function will generate q-q  plot and line
# given two vectors: a (y-axis) and b (x-axis)

q.q <- function(a,b, lin=TRUE){
  probs <-  c(0.25, 0.75)
  la <- length(a)
  lb <- length(b)
  a  <- sort(a)
  b  <- sort(b)
  fa <- ( 1:la - 0.5) / la
  fb <- ( 1:lb - 0.5) / lb
  if (la < lb) {
    b <- approx(fb, b, fa)$y 
  } else if( la > lb) {
    a <- approx(fa, a, fb)$y
  } else{}
  y <- quantile(a,probs)
  x <- quantile(b,probs)
  slope <- diff(y)/diff(x)
  int <- y[1] - slope * x[1]
  plot(a ~ b, cex=0.7, pch=20, col="blue")
  if (lin == TRUE) {
    abline(int,slope)
  }
}

# Set sample size and compute uniform
# values

n  <- 1000 # Number of simulated samples
fi <- (1:n - 0.5)/n
b.shp1 <- c(1, 5 , 50, 10, 10)
b.shp2 <- c(10,  10 , 70, 5, 1)

# Normal q-q plot
b  <- qnorm(fi)

OP <- par( mfcol = c(2,5), mar = c(2,2,1,1) )
for (i in 1:5 ) {
  a <- qbeta(fi,shape1=b.shp1[i], shape2 = b.shp2[i])
  plot(density(a),main=NA,xlab=NA)
  q.q(a,b)
}
par(OP)

```

## The `eda_qq` function

You were introduced to the `eda_qq` custom function in the previous chapter. This function can also be used to generate normal q-q plots by setting `norm` to `TRUE`. In such a case, the function takes as input a single vector of values.

```{r  class.source="eda", fig.height=3, fig.width=2.5, results='hide'}
library(tukeyedar)
eda_qq(alto, norm=TRUE) 
```

The function defaults to the quantile type `5`. To adopt the default quantile type used in `ggplot2`, set `q.type = 7`.

Note that when the `eda_qq` function is used to generate a normal q-q plot, the light dashed lines highlight the standard deviation for both sets of values. This differs from the mid 75% representation of values adopted by the function when generating an empirical q-q plot.
