---
title: "Manipulating data tables with dplyr"
editor_options: 
  chunk_output_type: console
---

```{r echo=FALSE}
source("libs/Common.R")
options(width = 80)
```


<div style="color:#ff7535; background-color:#fff0ee ;	border-left-style: solid">
This tutorial makes use of the following R package(s): **`dplyr`**, **`lubridate`** and **`forcats`**.
</div>
 


The data file  *FAO_grains_NA.csv* will be used in this exercise. This dataset consists of grain yield and harvest year by North American country. The dataset was downloaded from http://faostat3.fao.org/ in June of 2014. 

Run the following line to load the FAO data file into your current R session.

```{r}
dat <- read.csv("http://mgimond.github.io/ES218/Data/FAO_grains_NA.csv", header=TRUE)
```


# The `dplyr` basics

The basic set of R tools can accomplish many data table queries, but the syntax can be overwhelming and verbose. The package `dplyr` offers some nifty and simple querying functions as shown in the next subsections. Some of `dplyr`'s key data manipulation functions are summarized in the following table:

`dplyr` function   | Description
-------------------|-------------
`filter()`         | Subset by row values
`arrange()`        | Sort rows by column values
`select()`         | Subset columns
`mutate()`         | Add columns
`summarise()`      | Summarize columns

Note that all of these functions take as first argument the data table name except when used in a piping operation (pipes will be discussed later in this section). For example:

```{r eval=FALSE}
# When used alone, dataframe dat is inside the select function
dat2 <- select(dat, column1)

# When used in a pipe, dataframe dat is outside of the select function
dat2 <- dat %>% select(column1)
```

Make sure to load the `dplyr` package before proceeding with the following examples. 

```{r, message=FALSE}
library(dplyr)
```

## Subset by rows: `filter`

Tables can be subsetted by rows based on column values. For example, we may wish to grab all rows associated with `Oats`: 

```{r}
dat.query1 <- filter(dat, Crop == "Oats")
summary(dat.query1)
```

Note that R is case sensitive, so make sure that you respect each letter's case (i.e. upper or lower).

We can expand our query by including both `Oats`, `Buckwheat` and limiting the country to `Canada`.

```{r}
dat.query2 <- filter(dat, Crop == "Oats" | Crop == "Buckwheat", 
                          Country == "Canada")
summary(dat.query2)
```

The character `|` is the Boolean operator OR. So in our example, the query can be read as "... crop equals oats OR crop equals buckwheat". Had we used the AND operator, `&`, instead as in `Crop == "Oats" & Crop == "Buckwheat"` the output would have returned zero rows since a `Crop` value cannot be both `Oats` AND `Buckwheat`.

We can expand this query by limiting our output to the years 2005 to 2010

```{r}
library(dplyr)
dat.query3 <- filter(dat, Crop == "Oats" | Crop == "Buckwheat", 
                          Country == "Canada", 
                          Year >= 2005 & Year <= 2010)
summary(dat.query3)
```

Note the use of the AND Boolean operator (`&`) instead of the OR operator (`|`) for the `Year` query. We want the `Year` value to satisfy two criteria simultaneously: *greater than or equal to 2005* **AND** *less than or equal to 2010*. Had we used the `|` operator, R would have returned all years since all year values satisfy at least one of the two criterion.

## Sort rows by column value: `arrange`

You can sort a table based on a column's values. For example, to sort `dat` by crop name type:

```{r}
dat.sort1 <- arrange(dat, Crop)
head(dat.sort1)
tail(dat.sort1)
```

By default, `arrange` sorts by ascending order. To sort by descending order, wrap the column name with the function `desc()`. For example, to sort the table by `Crop` in ascending order then by `Year` in descending order, type:

```{r}
dat.sort2 <- arrange(dat, Crop, desc(Year))
head(dat.sort2)
tail(dat.sort2)
```

## Subset by column: `select`

You can grab a subset of the table by column(s). To extract the columns `Crop`, `Year` and `Value`, type:

```{r}
dat.subcol <- select(dat, Crop, Year, Value)
head(dat.subcol, 2)
```

If you want all columns *other* than `Crop`, `Year` and `Value`, add the negative `-` symbol before the column name:

```{r}
dat.subcol <- select(dat, -Crop, -Year, -Value)
head(dat.subcol, 2)
```

## Creating and/or calculating column values: `mutate`

You can add columns (and compute their values) using the `mutate` function. For example, to add a column `Ctr_abbr` and assign it the abbreviated values `CAN` for Canada and `USA` for the United States of America based on the values in column `Country` type:

```{r}
dat.extended <- mutate(dat, Ctr_abbr = ifelse(Country == "Canada", "CAN", "USA"))
head(dat.extended,3)
tail(dat.extended,3)
```

Here, we make use of an embedded function, `ifelse`, which performs a conditional operation: if the `Country` value is `Canada` return `CAN` if not, return `USA`.

Note that if you wish to *rename* a column, you can use the `rename()` function instead of `mutate`.

You can also use `mutate` to recompute column values. For example, to replace the `Country` column values with `CAN` or `USA` type:

```{r}
dat.overwrite <- mutate(dat, Country = ifelse(Country == "Canada", "CAN", "USA"))
head(dat.overwrite,3)
tail(dat.overwrite,3)
```

## Creating and/or calculating values across muliple columns: `mutate_at`

You might find yourself wanting to apply a same set of *mutate* operations across multiple variables. For example, given the following sample dataset,

```{r}
met <- data.frame(Wind = c(3.4, 5.0, 99, 4.1, 1.5),
                  Dir  = c(181, 220, 15,  15,  99 ),
                  Prec = c(99 , 0.5,  0,  99,  99))
met
```

where the value `99` is a placeholder for a missing value for the variables `Wind` and `Prec` but a valid value for `Dir`, we want to replace all missing values with `NA`. We could either create two mutate operations as in,

```{r}
met2 <- mutate( met, Wind = ifelse(Wind == 99, NA, Wind),
                     Prec = ifelse(Prec == 99, NA, Prec))
met2
```

or, we could reduce the separate `mutate` operations into a single `mutate_at` operation,

```{r}
met2 <- mutate_at( met, vars(Wind, Prec),
                        funs(ifelse( . == 99, NA, .)))
met2
```
The dots `.` in the `funs()` function are placeholders for each column listed in the `vars()` function.

## Summarize columns: `summarise`

You can summarize (or "collapse") one or more columns using the `summarise` function. For instance, to get the minimum and maximum years from the `Year` column, type:

```{r}
summarise(dat, yr_min = min(Year), yr_max=max(Year))
```

# Combining table manipulation functions using `%>%`

In most cases, you will find yourself wanting to combine several of `dplyr`'s data manipulation functions. For example, 

```{r}
dat.yield  <- filter(dat, Information == "Yield (Hg/Ha)", 
                          Crop == "Oats",
                          Year == 2012)
dat.rename <- mutate(dat.yield, Country = ifelse(Country == "Canada", "CAN", "USA"))
dat.final  <- select(dat.rename, Country, Value)

head(dat.final, 3)
```

The downside to this approach is the creation of several intermediate objects (e.g. `dat.yield` and `dat.rename`). This can make the workflow difficult to follow and clutter your R session with needless intermediate objects. 

Another approach to combining `dplyr` operations is to use the **piping operator** ,`%>%`, which daisy chains `dplyr` operations. So our previous workflow could look like:

```{r}
dat.final <- dat %>%
  filter(Information == "Yield (Hg/Ha)", 
         Crop == "Oats",
         Year == 2012)  %>% 
  mutate(Country = ifelse(Country == "Canada", "CAN", "USA")) %>%
  select(Country, Value)

head(dat.final, 3)
```

The chunk of code can be read as *"... with the `dat` table, `filter` by ..., then `mutate` ...., then `select` ..."* with the result from one operation being passed on to the next using the `%>%` operator. Note that the `filter`, `mutate` and `select` functions do not include the data table name making the chunk of code less cluttered and easier to read. The input data table  `dat` appears just once at the beginning of the pipe.


# Conditional statements

## The base `ifelse`

Conditional statements are used when you want to create an output value that is conditioned on an evaluation. For example, if you want to output a value of `1` *if* an input value is less than `23` and a value of `0` otherwise, you can make use of the `ifelse` function as follows:

```{r}
x <- c(12,102, 43, 20, 90, 0, 12, 6)
ifelse(x < 23, 1, 0)
```

The  base `ifelse` function works as expected when the input/output values are  numeric or character, but does not work as expected when applied to factors or dates. For example, if you wish to replace one factor level with another, the following example will not return the expected output.

```{r}
x <- as.factor( c("apple", "banana", "banana", "pear", "apple"))
ifelse(x == "pear", "apple", x)
```

The output is a character representation of the level number (recall that factors encode level values as numbers behind the scenes, i.e. `apple` =1, `banana`=2, etc...). Likewise, if you wish to replace an erroneous date with a missing value you will get:

```{r}
library(lubridate)
y <- mdy("1/23/2016", "3/2/2016", "12/1/1901", "11/23/2016")
ifelse( year(y) != 2016, NA, y)
```

Here, `ifelse` converts the date object to its internal numeric representation as number of days since 1970. 

If you want to ensure that the data type is preserved, a safer alternative is to use `dplyr`'s `if_else` function.

## `dplyr`'s `if_else`

The `if_else` function (note the underscore `_`), will preserve data type but does so strictly. For example, the following code will return an error: 

```{r error=TRUE}
if_else( year(y) != 2016, NA, y)
```

The output data types (`NA` and `y`) are not treated the same (by seeing `NA`, the function assumes that the output must be logical which `y` is not). The solution is to force `NA` as a *date* object by wrapping it with a date function like `mdy(NA)`:

```{r}
if_else( year(y) != 2016, mdy(NA), y)
```

Likewise, if the input vector is of type character, you need to ensure that all output values are characters too.

```{r}
y <- c("apple", "banana", "banana", "pear", "apple")
if_else( y == "banana", as.character(NA), y)
```

Recall from the Week 02 lecture notes that R has several `NA` reserved words for different data types (e.g. `NA_character_` and `NA_integer_` to name a few). So the last chunk of code could have been written as,

```{r}
if_else( y == "banana", NA_character_, y)
```
Note that there is no reserved word for `NA` *date* types in which case you would coerce `NA` to date using built-in functions like `as.Date(NA)`  or lubridate functions like `mdy(NA)` as demonstrated in an earlier example.


## Recoding factors using `recode`

When working with factors, however,  `if_else` (as of `dplyr` version `0.7`) will produce the following error:

```{r error=TRUE}
if_else(x == "pear", "apple", x)
```

R sees `apple` as a character and `x` as a number (i.e. the numeric representation of the factor level). A better option for recoding factors is to use `dplyr`'s `recode` function:


```{r}
recode(x , "pear" = "apple")
```

You can recode more than one factor level. In fact, you can even introduce new levels in the recoding scheme:

```{r}
recode(x , "pear" = "apple", "banana" = "pineapple" )
```

Note that as with `if_else` and `case_when`, `recode` is strict about preserving data types. So if you want to recode a level to `NA` make sure to use `NA_character_` or `as.character(NA)`.

### Replacing NA factor levels

One operation you cannot perform with `recode` is converting an `NA` level to another level. For example, the following will generate an error message:

```{r}
x[2] <- NA
x
```

```{r eval=FALSE}
recode(x, NA = "other")
```

```
Error: unexpected '=' in "recode(x, NA ="
```

The simplest solution is to make use of a specialized factor package called `forcats` and its function, `fct_explicit_na`.

```{r}
library(forcats)
x <- fct_explicit_na(x, na_level = "Other")
x
```

## Changing values based on multiple conditions: `case_when`

`ifelse` and `if_else` work great when a single set of conditions is to be satisfied. But if multiple sets of conditions are to be tested, nested if/else statements become cumbersome and are prone to clerical error. The following code highlights an example of nested if/else statements.

```{r}
z <- c("deg","F","C", "Temperature")
if_else( z == "deg", "Degree", if_else(z == "F", "Fahrenheit", "Celsius"))
```

A better alternative is to use `case_when`.

```{r}
case_when(z == "deg" ~ "Degree",
          z == "F" ~ "Fahrenheit",
          z == "C" ~ "Celsius",
          TRUE ~ z)
```
The last parameter, `TRUE ~ z` instructs R to keep all other values as is (not doing so would return `NA` values).

As with the `if_else` function, `case_when` is strict about data type, so if the input is a character vector it expects the output to be character as well. So if you are to assign `NA` values in your output, avoid the following mistake.

```{r error=TRUE}
case_when(z == "deg" ~ NA,  # NA is a logical type, this wil return an error
          z == "F" ~ "Fahrenheit",
          z == "C" ~ "Celsius",
          TRUE ~ z)
```
As was the case with the earlier `if_else` example, `case_when` requires a character `NA` type. We can use either `as.character(NA)` or `NA_character_`.

```{r}
case_when(z == "deg" ~ NA_character_,
          z == "F" ~ "Fahrenheit",
          z == "C" ~ "Celsius",
          TRUE ~ z)
```

Note that `ifelse`, `if_else`, `recode` and `case_when` can all be used inside of a `mutate` function. For example, to replace `Canada` and `United States of America` in variable `Country` with `CAN` and `USA` respectively and to create a new variable called `Type` which will take on the values of `1`, `2` or `3` depending on the values in variable `Source`, type the following: 

```{r}
dat1 <- dat %>% 
  mutate(Country = recode(Country, "Canada" = "CAN",
                                   "United States of America" = "USA"),
         Type = case_when(Source == "Calculated data" ~ 1,
                          Source == "Official data" ~ 2,
                          TRUE ~ 3)) 
head(dat1)   
```

# Miscellaneous

## Outputting a vector instead of a table using `pull`

Piping operations will normally output a table, even if a single value is returned. For example, the following summarization operation returns the total oats yield as a data table:

```{r}
oats <- dat %>% 
  filter(Crop == "Oats",
         Information == "Yield (Hg/Ha)") %>% 
  summarise(Oats_sum = sum(Value))
oats
class(oats)
```

There may be times when you want the output as a vector element and not a data table. To output a vector, use the `pull()` function.

```{r}
oats <- dat %>% 
  filter(Crop == "Oats",
         Information == "Yield (Hg/Ha)") %>% 
  summarise(Oats_sum = sum(Value)) %>% 
  pull()
oats

class(oats)
```

The `pull` function can also be used to convert a data table column to a multi-element vector, e.g.:

```{r}
# This outputs a one column table
yield <- dat %>% 
  filter(Crop == "Oats",
         Information == "Yield (Hg/Ha)") %>% 
  select(Value)

head(yield)

class(yield)
```

```{r}
# This outputs a multi-element vector
yield <- dat %>% 
  filter(Crop == "Oats",
         Information == "Yield (Hg/Ha)") %>% 
  pull(Value)

head(yield)

class(yield)
```
