---
title: "Bivariate analysis"
---

```{r echo=FALSE}
source("libs/Common.R")
```

-----
```{r echo = FALSE}
pkg_ver(c("ggplot2"))
```
-----

<em>
This material is intended to supplement **pages 87 to 105** of **Cleveland's book**.
</em>

Bivariate data are datasets that store two variables measured from a same observation (e.g. wind speed **and** temperature at a single location). This differs from univariate data where only one variable is measured for each observation (e.g. temperature at a single location). 

# Scatter plot

A scatter plot is a popular visualization tool used to compare values between two variables. Sometimes one variable is deemed *dependent* on another variable; the latter being the *independent* variable. Cleveland refers to the former as the *response*  and the latter as the *factor* (this is not to be confused with the *factor* data type used in R as a grouping variable). The dependent variable is usually plotted on the y-axis and the independent variable is usually plotted on the x-axis. Other times, one does not seek a dependent-independent relationship between variables but is simply interested in studying the relationship between them.

A scatter plot can be generated using the base plotting environment as follows:

```{r fig.width=2.5, fig.height=2.5, small.mar=TRUE}
df <- read.csv("http://mgimond.github.io/ES218/Data/ganglion.csv")

plot(cp.ratio ~ area, dat = df)
```

Or, using `ggplot2`, as follows:

```{r fig.width=2.5, fig.height=2.5}
library(ggplot2)
ggplot(df, aes(x = area, y = cp.ratio)) + geom_point()
```

The data represent the ratio between the ganglion cell density of a cat's central retina to that of its peripheral density (variable `cp.ratio`) and the cat's retina surface area (`area`) during its early development (ranging from 35 to 62 days of gestation).

# Fitting the data

Scatter plots are a good first start in visualizing the data, but this is sometimes not enough. Our eyes need "guidance" to help perceive patterns. Another visual aid involves **fitting** the data with a line. 

## Parametric fit

### Fitting a straight line

A straight line is the simplest fit one can make to bivariate data. A popular method for fitting a straight line is the  least-squares method. We'll use R's `lm()` function which provides us with a slope and intercept for the best fit line.

In the base plotting environment, we can do the following:

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
M <- lm(cp.ratio ~ area, dat = df)
plot(cp.ratio ~ area, dat = df)
abline(M, col = "red")
```

In the `ggplot2` plotting environment, we can make use of the `stat_smooth` function to generate the regression line.

```{r fig.width=2.5, fig.height=2.5}
library(ggplot2)
ggplot(df, aes(x = area, y = cp.ratio)) + geom_point() + 
             stat_smooth(method ="lm", se = FALSE)
```

The `se = FALSE` option prevents R from drawing a confidence envelope around the regression line.

The straight line is a **first order polynomial** with two parameters, $a$ and $b$, that define an equation that best describes the relationship between the two variables:

$$
CP_{Ratio} = a + b (Area)
$$

where $a$ and $b$ can be extracted from the regression model object `M` as follows:

```{r}
coef(M)
```

Thus $a$ = 0.014 and $b$ = 0.11.

### Fitting a 2^nd^ order polynomial

A second order polynomial is a three parameter function ($a$, $b$ and $c$) whose equation $y = a + bx + cx^2$ defines the curve that best fits the data. We define such a relationship in R using the formula `cp.ratio ~ area + I(area^2)`. The *identity* function `I()` preserves the arithmetic interpretation of `area^2` as part of the model. Our new `lm` expression and resulting coefficients follow:

```{r}
M2 <- lm(cp.ratio ~  area + I(area^2) , dat = df)
coef(M2)
```

The quadratic fit is thus,

$$
y = 2.87 - 0.012 x + 0.000839 x^2
$$

We cannot use `abline` to plot the predicted 2^nd^ order polynomial curve since `abline` only draws straight lines. We will need to construct the line manually using the `predict` and `lines` functions.

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
plot(cp.ratio ~ area, dat=df)
x.pred <- data.frame( area = seq(min(df$area), max(df$area), length.out = 50) )
y.pred <- predict(M2, x.pred)
lines(x.pred$area, y.pred, col = "red")
```

To do this in `ggplot2` simply pass the formula as an argument to `stat_smooth`:

```{r fig.width=2.5, fig.height=2.5}
ggplot(df, aes(x = area, y = cp.ratio)) + geom_point() + 
  stat_smooth(method = "lm", se = FALSE, formula = y ~  x + I(x^2) )
```


## Non-parametric fits

### Loess

A more flexible curve fitting option is the **loess**  curve (short for **lo**cal regr**ess**ion; also known as the *local weighted regression*). Unlike the parametric approach to fitting a curve, the loess does **not** impose a structure on the data. The loess curve fits small segments of a regression lines across the range of x-values, then links the mid-points of these regression lines to generate the smooth curve. The range of x-values that contribute to each localized regression lines is defined by the $\alpha$ parameter which usually ranges from 0.2 to 1. The larger the $\alpha$ value, the smoother the curve. The other parameter that defines a loess curve is $\lambda$: it defines the polynomial order of the localized regression line. This is usually set to 1 (though `ggplot2`'s implementation of the loess defaults to a 2^nd^ order polynomial).

```{r echo=FALSE}
library(dplyr)
library(purrr)

alpha <-  0.5  
strip.x <- nrow(df) * alpha # Number of points within band

f.plot <- function(start , line = FALSE, point = FALSE, bnds =FALSE,
                   w = FALSE, plot = TRUE, pts = FALSE, title = NULL){
  # Find points closest to starting point
  subset <-  df %>% 
    mutate(dst = abs(area - start)) %>% 
    arrange(dst) %>% 
    mutate(j = row_number()) %>% 
    filter(j <= strip.x ) %>% 
    mutate(wt = dst / max(dst) * 3)
  
  # Assign weights
  wts <- dnorm(subset$wt)/ 0.3989423
  
  # Regress with weights
  M <- lm(cp.ratio ~ area, subset, weights = wts)
  x.l <- coef(M)[1] + coef(M)[2] * start
  
  # Plot by option
  if(plot == TRUE){
    plot(cp.ratio ~ area, df,  yaxt='n', main = title,
         axes = FALSE, pch=16, col = "grey90", cex = 1.6)
    axis(side=1, at=c(seq(10,150,30)))
    abline(v = start, lty = 2)
    if(bnds == TRUE){
      abline(v = c(min(subset$area),max(subset$area)), lty = 3, col = "grey")
      rect(min(subset$area), 0, max(subset$area), 20, col = rgb(0,0,1,0.1),
           border = rgb(0,0,1,0.1))
    }
    
    if(w == TRUE){
      points(x = subset$area, y = subset$cp.ratio, col = rgb(0,0,1, wts), 
             pch = 16,  cex=1.6)
    }
    if(line == TRUE){
      clip(min(subset$area),max(subset$area),
           min(df$cp.ratio),max(df$cp.ratio))
      abline(M, col = "orange", lwd = 1.8)
    }
    if(point == TRUE){
      points(x = start, y = x.l, pch = 16, col = "red", cex=1.8) 
    }
  }
  
  if(pts == TRUE){
    return(c(start, x.l))
  }  
  
}

```

### How a loess is constructed

Behind the scenes, each point (x~i~,y~i~) that defines the loess curve is constructed as follows: 
 
 a) A subset of data points closest to point x~i~ are identified. The number of points in the subset is computed by multiplying the bandwidth $\alpha$ by the total number of observations. In our current example, the number of points defining the subset is 0.5 * 14 = 7. The points are identified in the light blue area of the plot in panel (a) of the figure below.
 b) The points in the subset are assigned weights. Greater weight is assigned to points closest to x~i~ and vice versa. The weights define the points' influence on the fitted line. Different weighting techniques can be implemented in a loess with the `gaussian` weight being the most common. Another weighting strategy we will also explore later in this course is the `symmetric` weight.
 c) A regression line is fit to the subset of points. Points with smaller weights will have less leverage on the fitted line than points with larger weights. The fitted line can be either a first order polynomial fit or a second order polynomial fit.
 d) Compute the value y~i~ from the regression line. This is shown a the red dot in panel (d). This is one of the points that will define the shape of the loess.
   

```{r fig.height = 2.7, fig.width = 10, echo = FALSE}
OP <- par(mfrow = c(1, 4), mar=c(2,1,1,0), pty = "s")
f.plot(start = 50, bnds = TRUE, title = "(a) Subset points") 
f.plot(start = 50, bnds = TRUE, w = TRUE, title = "(b) Assign weights")
f.plot(start = 50, bnds = TRUE, w = TRUE, line = TRUE, title = "(c) Fit line") 
f.plot(start = 50,  bnds = TRUE, w = TRUE, line = TRUE, point = TRUE, 
       title = "(d) Draw point")
par(OP)
```

The above steps are repeated for as many x~i~ values practically possible. Note that when x~i~ approaches an x limit, the subset of points becomes skewed to one side of x~i~. For example, when estimating x~10~, the seven closest points to the right of x~10~ are selected.

```{r fig.height = 2.5, fig.width = 10, echo = FALSE}
OP <- par(mfrow = c(1, 4), mar=c(2,1,1,0), pty = "s")
f.plot(start = 10,  bnds = TRUE, w = TRUE, line = TRUE, point = TRUE, 
       title = expression(Left-most ~~  x[i] )) 
f.plot(start = 140,  bnds = TRUE, w = TRUE, line = TRUE, point = TRUE, 
       title = expression(Right-most ~~  x[i]) ) 
par(OP)
```


In the following example, just under 30 loess points are computed at equal intervals. This defines the shape of the loess.

```{r fig.height = 2.3, fig.width = 10, echo = FALSE}
OP <- par(mfrow = c(1, 4), mar=c(2,1,1,0), pty = "s")
l.pts <- seq(10,140,(110-20)/20) %>% map(function(x) f.plot(start = x, plot = FALSE, pts = TRUE)) %>% 
  do.call(rbind, .) %>% as.data.frame()
plot(cp.ratio ~ area, df,  yaxt='n', main = NULL,
         axes = FALSE, pch=16, col = "grey90", cex = 1.6)
    axis(side=1, at=c(seq(10,150,30)))
points(l.pts$V1, l.pts$`(Intercept)`, pch = 16, col = "red",
       cex=1)
par(OP)
```

It's more conventional to plot the line segments than it is to plot the points.

```{r fig.height = 2.3, fig.width = 10, echo = FALSE}
OP <- par(mfrow = c(1, 4), mar=c(2,1,1,0), pty = "s")
l.pts <- seq(10,140,(110-20)/20) %>% map(function(x) f.plot(start = x, plot = FALSE, pts = TRUE)) %>% 
  do.call(rbind, .) %>% as.data.frame()
plot(cp.ratio ~ area, df,  yaxt='n', main = NULL,
         axes = FALSE, pch=16, col = "grey90", cex = 1.6)
    axis(side=1, at=c(seq(10,150,30)))
lines(l.pts$V1, l.pts$`(Intercept)`, col = "red")
par(OP)
```

<br>

### Plotting a loess in R

The loess fit can be computed using the `loess()` function. It takes as arguments  `span` ($\alpha$), and `degree` ($\lambda$).

```{r}
# Fit loess function
lo <- loess(cp.ratio ~ area, df, span = 0.5, degree = 1)

# Predict loess values for a range of x-values
lo.x <- seq(min(df$area), max(df$area), length.out = 50)
lo.y <- predict(lo, lo.x)
```

The predicted loess curve can be added using the `lines` function.

```{r fig.width=2.5, fig.height=2.5, small.mar=TRUE}
plot(cp.ratio ~ area, dat = df)
lines(lo.x, lo.y, col = "red")
```

In `ggplot2` simply pass the `method="loess"` parameter to the `stat_smooth` function.

```{r eval=FALSE}
ggplot(df, aes(x = area, y = cp.ratio)) + geom_point() + 
             stat_smooth(method = "loess", se = FALSE, span = 0.5)
```

However, `ggplot` (up to version 3.3) defaults to a second degree loess (i.e. the small regression line elements that define the loess are defined by a 2^nd^ order polynomial and not a 1^st^ order polynomial). If a first order polynomial (`degree=1`) is desired, you need to include an argument list in the form of `method.args=list(degree=1)` to the `stat_smooth` function.

```{r fig.width=2.5, fig.height=2.5}
ggplot(df, aes(x = area, y = cp.ratio)) + geom_point() + 
             stat_smooth(method = "loess", se = FALSE, span = 0.5, 
                         method.args = list(degree = 1) )
```

# Residuals

Fitting the data with a line is just the first step in EDA. Your next step should be to explore the residuals. The residuals are the distances (along the y-axis) between the observed points and the fitted line. The closer the points to the line (i.e. the smaller the residuals) the better the fit. 

The residuals can be computed using the `residuals()` function. It takes as argument the model object. For example, to extract the residuals from the linear model `M` computed earlier type, 

```{r}
residuals(M)
```

## Residual-dependence plot

We'll create a **residual dependence plot** to plot the residuals as a function of the x-values. We'll do this using `ggplot` so that we can also fit a loess curve to help discern any pattern in the residuals (the `ggplot` function makes it easier to add a loess fit than the traditional plotting environment).


```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
df$residuals <- residuals(M)
ggplot(df, aes(x = area, y = residuals)) + geom_point() +
             stat_smooth(method = "loess", se = FALSE, span = 1, 
                         method.args = list(degree = 1) )

```

We are interested in identifying any pattern in the residuals. If the model does a good job in fitting the data, the points should be uniformly distributed across the plot and the loess fit should approximate a horizontal line. With the linear model `M`, we observe a convex pattern in the residuals suggesting that the linear model is not a good fit. We say that the residuals show *dependence* on the x values. 

Next, we'll look at the residuals from the second order polynomial model `M2`.

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
df$residuals2 <- residuals(M2)
ggplot(df, aes(x = area, y = residuals2)) + geom_point() +
               stat_smooth(method = "loess", se = FALSE, span = 1, 
                           method.args = list(degree = 1) )

```

There is no indication of dependency between the residual and the `area` values. The second order polynomial is an improvement over the first order polynomial. Let's look at the loess model.


```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
df$residuals3 <- residuals(lo)
ggplot(df, aes(x = area, y = residuals3)) + geom_point() +
               stat_smooth(method = "loess", se = FALSE, span = 1, 
                           method.args = list(degree = 1) )

```

The loess model also seems to do a good job in smoothing out any overall pattern in the data.

You may ask *“if the loess model does such a good job in fitting the data, why bother with polynomial fits?”* If you are seeking to generate a predictive model that explains the relationship between the y and x variables, then a mathematically tractable model (like a polynomial model) should be sought. If the interest is simply in identifying a pattern in the data, then a loess fit is a good choice.

Next we will look for homogeneity in the residuals.

## Spread-location plot

The `M2` and `lo` models do a good job in eliminating any dependence between residual and x-value. Next, we will check that the residuals do not show a dependence with *fitted* y-values. This is analogous to univariate analysis where we checked if residuals increased or decreased with increasing medians across factors. Here we will compare residuals to the fitted `cp.ratio` values (think of the fitted line as representing a *level* across different segments along the x-axis). We'll generate a spread-level plot of model `M2`'s residuals (note that in the realm of regression analysis, such plot is often referred to as a **scale-location** plot). We'll also add a loess curve to help visualize any patterns in the plot (this reproduces fig 3.14 in Cleveland's book).

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
sl2 <- data.frame( std.res = sqrt(abs(residuals(M2))), 
                   fit     = predict(M2))

ggplot(sl2, aes(x = fit, y  =std.res)) + geom_point() +
              stat_smooth(method = "loess", se = FALSE, span = 1, 
                          method.args = list(degree = 1) )
```

The function `predict()` extracts the y-values from the fitted model `M2` and is plotted along the x-axis. It's clear from this plot that the residuals are not homogeneous; they increase as a function of increasing *fitted* CP ratio. The "bend" observed in the loess curve is most likely due to a single point at the far (right) end of the fitted range. Given that we have a small batch of numbers, a loess can be easily influenced by an outlier. We may want to increase the span.


```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
ggplot(sl2, aes(x = fit, y = std.res)) + geom_point() +
              stat_smooth(method = "loess", se = FALSE, span = 1.5, 
                          method.args = list(degree = 1) )
```

The point's influence is reduced enough to convince us that the observed monotonic increase is real (Note that we would observe this monotone spread with our loess model as well). At this point, we should look into re-expressing the data.

## Checking residuals for normality

If you are interested in conducting a hypothesis test (i.e. addressing the question *"is the slope significantly different from 0"*) you will likely want to check the residuals for normality since this is an assumption made when computing a confidence interval and a p-value. We'll make use of `geom_qq` and `geom_qq_line` to compare the residuals to a normal distribution.

```{r fig.height=2.5, fig.width=2.5}
ggplot(df, aes(sample = residuals2)) + geom_qq(distribution = qnorm) +
  geom_qq_line(distribution = qnorm, col = "blue")
```

Here, the residuals seem to stray a little from a normal distribution.

# Re-expressing the data

The monotone spread can be problematic if we are to characterize the spread of `cp.ratio` as being the same across all values of `area`. To remedy this, we can re-express the `cp.ratio` values. Ratios are good candidates for log transformation. We will therefore fit a new linear model to the data after transforming the y-value.

```{r}
df.log <- data.frame( area = df$area, cp.ratio.log = log(df$cp.ratio))
M3     <- lm(cp.ratio.log ~ area, dat = df.log)
```

Next, let's plot the transformed data and add the fitted line.

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
ggplot(df.log, aes(x = area, y = cp.ratio.log)) + geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```
 
At first glance, the log transformation seems to have done a good job at straightening the batch of values. Next, let's look at the residual dependence plot.

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
df.log$residuals <- residuals(M3)
ggplot(df.log, aes(x = area, y = residuals)) + geom_point() +
               stat_smooth(method = "loess", se = FALSE, span = 1, 
                           method.args = list(degree = 1) )
```

Logging the y values has eliminated the residual's dependence on `area`. Next, let's assess homogeneity in the residuals using the s-l plot.

```{r fig.height=2.5, fig.width=2.5, small.mar=TRUE}
sl3 <- data.frame( std.res = sqrt(abs(residuals(M3))), 
                   fit     = predict(M3))
ggplot(sl3, aes(x = fit, y = std.res)) + geom_point() +
               stat_smooth(method ="loess", se = FALSE, span = 1, 
                           method.args=list(degree = 1) )
```

We do not observe a systematic increase in spread, the log transformation seems to have removed the monotone spread.

Finally, we'll check for normality of the residuals.

```{r fig.height=2.5, fig.width=2.5}
ggplot(df.log, aes(sample = residuals)) + geom_qq(distribution = qnorm) +
  geom_qq_line(distribution = qnorm, col = "blue")
```

An added benefit of re-expressing the data seems to be a slight improvement in the normality of the residuals.
