---
title: "Week3: Querying data tables"
output:
  html_document:
    toc: yes
    toc_depth: 3
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  comment = "",
  message = FALSE,
  tidy = FALSE,
#  dev=c('CairoPNG', 'CairoPDF'),
  type = "cairo",
  cache = FALSE,
  warning = FALSE,
  encoding = "UTF-8",
#  pointsize=12* 200 / 72, # = size * DPI / 72
  res=200)
```

The data file  *FAO_grains_NA.csv* will be used in this exercise. This dataset consists of grain yield and harvest year by north american country. The dataset was downloaded from http://faostat3.fao.org/ in June of 2014. 

Run the following line to load the FAO data file into your current R session.

```{r}
dat <- read.csv("http://mgimond.github.io/ES218/Data/FAO_grains_NA.csv", header=TRUE)
```


# Querying tables using `dplyr`

The basic set of R tools can accomplish many data table queries, but the syntax can be overwhelming and verbose. The package `dplyr` offers some nifty and simple querying functions as shown in the next subsections. `dplyr`'s data manipulation functions are summarized in the following table:

`dplyr` function   | Description
-------------------|-------------
`filter()`         | Subset by row values
`arrange()`        | Sort rows by column values
`select()`         | Subset columns
`mutate()`         | Add columns
`summarise()`      | Summarize columns

Make sure to load the `dplyr` package before proceeding with the following examples. All examples assume that `dplyr` version 0.3 or greater is installed.

```{r, message=FALSE}
library(dplyr)
```

## Subset by rows

Tables can be subsetted by rows based on column values. For example, we may wish to grab all rows associated with `Oats`: 

```{r}
dat.query1 <- filter(dat, Crop == "Oats")
summary(dat.query1)
```

Note that R is case sensitive, so make sure that you respect each letter's case (i.e. upper or lower).

We can expand our query by including both `Oats`, `Buckwheat` and limiting the country to `Canada`.

```{r}
dat.query2 <- filter(dat, Crop == "Oats" | Crop == "Buckwheat", Country == "Canada")
summary(dat.query2)
```

The character `|` is the Boolean operator OR. So in our example, the query can be read as "... crop equals oats OR crop equals buckwheat". Had we used the AND operator, `&`, instead as in `Crop == "Oats" & Crop == "Buckwheat"` the output would have returned zero rows since a `Crop` value cannot be both `Oats` AND `Buckwheat`.

We can expand this query by limiting our output to the years 2005 to 2010

```{r}
library(dplyr)
dat.query3 <- filter(dat, Crop == "Oats" | Crop == "Buckwheat", Country == "Canada", Year >= 2005 & Year <= 2010)
summary(dat.query3)
```

Note the use of the Boolean operator AND, `&`, instead of OR for the `Year` query. We want the `Year` value to satisfy two criteria simultaneously: *greater than or equal to 2005* **AND** *less then or equal to 2010*. Had we used the `|` operator, R would have returned all years since all year values satisfy at least one of the two criteria.

## Sort rows by column value

You can sort a table based on a column's values. For example, to sort `dat` by crop name type:

```{r}
dat.sort1 <- arrange(dat, Crop)
head(dat.sort1)
tail(dat.sort1)
```

By default, `arrange` sorts by ascending order. To sort by descending order, wrap the column name with the function `desc()`. For example, to sort the table by `Crop` in ascending order then by `Year` in descending order, type:

```{r}
dat.sort2 <- arrange(dat, Crop, desc(Year))
head(dat.sort2)
tail(dat.sort2)
```

## Subset by column

You can grab a subset of the table by column(s). To extract the columns `Crop`, `Year` and `Value`, type:

```{r}
dat.subcol <- select(dat, Crop, Year, Value)
head(dat.subcol)
```

## Adding columns

You can add columns (and compute their values) using the `mutate` function. For example, to add a column `Ctr_abbr` and assign it the value `CAN` for Canada and `USA` for the United States of America based on the values in column `Country` type:

```{r}
dat.extended <- mutate(dat, Ctr_abbr = ifelse(Country == "Canada", "CAN", "USA"))
head(dat.extended)
tail(dat.extended)
```

## Summarize columns

You can summarize (or "collapse") one or more columns using the `summarise` function. For instance, to get the minimum and maximum years from the `Year` column, type:

```{r}
summarise(dat, yr_min = min(Year), yr_max=max(Year))
```

## Combining table manipulation functions using `%>%`

In most cases, you will find yourself wanting to combine several of `dplyr`'s data manipulation functions. For example, to summarize the total yield of Oats by country, you could type the following:

```{r}
dat.yield <- filter(dat, Information == "Yield (Hg/Ha)", Crop == "Oats")
dat.grp   <- group_by(dat.yield, Country)
dat.sum   <- summarise(dat.grp, mean_yield = mean(Value))
dat.sum
```

This block of code introduces you to another function called `group_by`. This function creates a new type of object (a list) which breaks the data into different groups for subsequent analysis. This implies that all functions that include `dat.grp` as input will output an object of similar class. So in our example, `dat.sum` is a list object and not a data frame. To force the `dat.sum` object back to a data frame, simply wrap it with the `as.data.frame()` function.

```{r}
as.data.frame(dat.sum)
```

The one downside to this approach of combining data table manipulation functions for complex queries is the use of several intermediate objects (e.g. `dat.yield`, `dat.grp`, `dat.sum`). This can make the workflow difficult to follow and clutter your R session with needless intermediate objects. 

Another approach to combining `dplyr` operations is to use the operator `%>%` which daisy chains `dplyr` functions. So our previous workflow would look like:

```{r}
dat %>%
  filter(Information == "Yield (Hg/Ha)", Crop == "Oats") %>%
  group_by(Country) %>%
  summarise(mean_yield = mean(Value))
```

The chunk of code can be read as "with the `dat` table, `filter` by ..., then `group_by` ...., then `summarise` by ..." with the result from one operation being passed on to the next using the `%>%` operator. Note that the `filter`, `group_by` and `summarise` functions do not include the data table name making the chunk of code less cluttered and easier to read.

This chunk of code will output the results to the screen, if you want to save it to an object called `dat2`, replace the first line with `dat2 <- dat %>%`

Here's a more complicated output whereby the median yield for each USA crop is summarized by year for the years spanning 2005 through 2010 and the output is sorted by crop name (in ascending order) and median yield value (in descending order).

```{r}
dat %>%
  filter(Information == "Yield (Hg/Ha)", Year >= 2005, Year <=2010, 
         Country=="United States of America") %>%
  group_by(Crop, Year) %>%
  summarise(median_yield = round(median(Value))) %>%
  arrange(Crop, desc(median_yield))
```

