---
title: "Week2: Data Object Type and Structure"
output:
  html_document:
    toc: yes
    toc_depth: 3
  word_document: default
---

# How to represent data in R

An object in R is a placeholder for different **types** of data and functions created or manipulated in R. The following sections focus on data **types** and then on data **objects**.

## Data types

R objects can store values as different data types including numeric, character, logical or date. R variable types are often referred to as **modes** and can be identified using the `mode()` function.

### Numeric

The numerical data type is probably the simplest. It consists of numbers such as integers (e.g. `1 ,-3 ,33 ,0`) or doubles (e.g. `0.3, 12.4, -0.04, 1.0`). For example, to create a numeric (double) vector we can type:

```{r}
x <- c(1.0, -3.4, 2, 140.1)
mode(x)
```

### Character

The character data type consists of letters or words such as `"a", "f", "project", "house value"`.

```{r}
x <- c("a", "f", "project", "house value")
mode(x)
```

Characters can also consist of numbers represented as characters. The distinction between a numeric representation of a number and a numeric one is important. For example, if we have two numeric vector `x` and `y` such as

```{r}
x <- 3
y <- 5.3
```

and we choose to sum the two variables, we get:

```{r}
x + y
```

If we repeat these steps but instead choose to represent the numbers `3` and `5.3` as characters we get the following error message:

```{r error=TRUE}
x <- "3"
y <- "5.3"
x + y
```

Note that if numbers are to be stored as characters, don't forget to wrap them around quotes.

```{r}
x <- c("1.0", "-3.4", "2", "140.1")
mode(x)
```

### Logical

Logical values can take on one of two values: `TRUE` or `FALSE`. These can also be represented as `1` or `0`. For example, to create a logical vector of 4 elements, you can type

```{r}
x <- c(TRUE, FALSE, FALSE, TRUE)
```

or
```{r}
x <- as.logical(c(1,0,0,1))
```

Note that in both cases, `mode(x)` returns `r mode(x)`. Also note that the `1`'s and `0`'s are converted to `TRUE`'s and `FALSE`'s.

### Factors

Factors are usually used to group variables into categories having a fixed number of levels. For example, a dataset may be grouped by gender or month of the year. Such data may be loaded into R as a numeric or character data type in which case it needs to be converted to a factor using the `as.factor()` function.

```{r}
a      <- c("M", "F", "F", "M", "F", "M", "M", "F", "F", "M")
a.fact <- as.factor(a)
x      <- c(166, 47, 61, 148, 62, 123, 232, 98, 93, 55)
dat    <- data.frame(x = x, gender = a.fact)
```

Many functions recognize factor data types and will allow you to split the output into groups. For example, to create a boxplot of the value `x` split into two groups, `F` and `M`, you can type the following:

```{r fig.width = 4, fig.height=2.5,echo=2}
OP <- par( mar=c(2,2,0,0))
boxplot(x ~ gender, dat, horizontal = TRUE)
par(OP)
```

### Dates

Date values can be represented in tables as numbers or characters. But to work properly with date representation, it must be converted to an R date object. R provides many facilities to convert and manipulate dates and times, but a package called `lubridate` makes working with dates/times much easier. The following subsections make use of the `lubridate` package.

## Data structure

### Vectors

Vector objects can be broken down into two types: atomic vector and lists.

### Atomic vectors

The *vector* is the simplest data structure in R which consists of an ordered set of values of the same type (e.g. numeric, character, date, etc...). The last section presented two examples of a vector object (`x <- 3.14` and `x <- runif(30,0,10)`). Another way of creating a vector object is by using the *combine* function `c()` as in

```{r}
x <- c(674 , 4186 , 5308 , 5083 , 6140 , 6381)
x
```

A vector object is an indexable collection of values and can thus allow one to access a specific index number. For example, to access the third element of `x`, type

```{r}
x[3]
```

You can also select a subset of elements by index values using the combine function `c()`,

```{r}
x[c(1,3,4)]
```

If you are interested in a range of indexed values, use `:`

```{r}
x[2:4]
```

You can also assign new values to a specific index. For example, we can replace the second value with 0

```{r}
x[2] <- 0
x
```

Note that a vector can also store characters or strings

```{r}
x <- c("all", "b", "olive")
x
```

### Lists

Like atomic vectors, a list is an ordered set of elements but unlike vectors, elements of a list can be of any type. A list is constructed using the `list()` function.

For example, the following list consists of 3 elements: a three element numeric vector(element `a`), a two element logical vector (element `b`) and a three element character vector (element `c`).

```{r}
l <- list(a = c(123.1, 3, 87.6), b = c(TRUE, FALSE), c = c("apples", "oranges", "round"))
```

You can view the contents of a list using the `str()` function.

```{r}
str(l)
```

Lists can also be recursive (i.e. a list can contain other lists). For example, the following list `l2` houses two elements: a two-by-two matrix (element `a`) and the list `l` (element `b`).

```{r}
l2 <- list( a = matrix( c(12.3, 20, 1.2, 18.0) , nrow=2), b = l)
str(l2)
```

You'll find that many functions in R return list objects such as the linear model function `lm`. For example, run a regression between two variables and save the output of the regression analysis to an object called `M`:

```{r}
x <- c(7.3, 29.4, 29.4, 2.9, 12.3, 7.5, 36.0, 4.8, 18.8, 4.2)
y <- c(5.2, 26.6, 31.2, 2.2, 13.8, 7.8, 35.2, 8.6, 20.3, 1.1)
M <- lm( y ~ x)
```

Now use the `typeof` function to view `M`'s data type:
```{r}
typeof(M)
```

The contents of the regression model object `M` consists of `r length(M)` elements which include various diagnostic statistics, regression coefficients, and residuals. 
```{r}
str(M)
```

Elements of a list can be extracted using the `$` symbol. For example, to access the values from `M`'s `residuals` element, type

```{r}
M$residuals
```

The element `residual` is a numeric vector. We can also access the contents of a recursive list element such as `qr`:

```{r}
M$qr
```

Note that since the element `qr` is a list, `M$qr` returns a list (and not a numeric vector as in the call to `M$residuals`). To extract the element `qraux` from the list `qr` (which is itself an element of `M`), you can use a recursive `$` as in:

```{r}
M$qr$qraux
```

Note that in the previous examples, a name was assigned to each element of the list (e.g. `a = matrix( c(12.3, 20, 1.2, 18.0)`) which allows us to access that element using the hash name (e.g. `l2$a`). We can list the element names of a list using the `names()` function:

```{r}
names(l2)
```

But lists can also be created without assigning names to an element. For example, list `l2` could be created as follows:

```{r}
l2.nonames <- list( matrix( c(12.3, 20, 1.2, 18.0) , nrow=2), l)
```

This last object has no names associated with its elements,

```{r}
names(l2.nonames)
```

This makes it impossible to access its elements using the `$` symbol. We must instead use the `[]` symbol to access the ordered elements. For example, to access the first element of `l2.nonames`, type:

```{r}
l2.nonames[1]
```

This displays the contents of the matrix, but the output is still a list which means that we cannot operate on the content as though it was a matrix (however, had the element been named as with the object `l2`, then the use of the `$` symbol would have returned a matrix as in `l2$a`).



### Matrices and arrays

Matrices in R can be thought of as vectors indexed using *two* indices. For example, the following line of code creates a 3 by 3 matrix of randomly generated values. The parameters `nrow` and `ncol` define the matrix 2D dimension and the function `runif()` generates the nine random numbers that populate this matrix.

```{r}
m <- matrix(runif(9,0,10), nrow = 3, ncol = 3)
m
```

If a higher dimension vector is desired, then use the `array()` function to generate the n-dimensional object. A 3x3x3 array can be created as follows:

```{r}
m <- array(runif(27,0,10), c(3,3,3))
m
```


Matrices and arrays can store numeric or character data types, but they cannot store both in a same object. This is not to say that you can't have a matrix of the kind

```{r echo=FALSE}
matrix(c("a","b",2,4), nrow=2)
```

but the value `2` and `4` are no longer treated as numeric values but as character values instead (as indicated by the quotes surrounding the numbers).

### Data frames

A *data frame* is what comes closest to our perception of a data *table*. It's an extension of the matrix object in that, unlike a matrix, a data frame can *mix* data types (e.g. both numeric and character data can coexist in a data frame). This will be the most frequently used data structure in this course with atomic vectors following closely behind.


```{r}
Name <- c("a1", "a2", "b3")
Value1 <- c(23, 4, 12)
Value2 <- c(1,45,5)
dat    <- data.frame(Name, Value1, Value2)
dat
```

Like a vector, elements of a data frame can be accessed by their index (aka subscripts). For example, to list the first two columns,

```{r}
dat[ , 1:2 ]
```

or the third row,

```{r}
dat[ 3 , ]
```

or the first entry of the data frame,

```{r}
dat[ 1 , 1 ]
```

or the values in the column named *Value2*,

```{r}
dat$Value2
```

