---
title: "Working with Dates"
editor_options: 
  chunk_output_type: console
---

```{r echo=FALSE}
source("libs/Common.R")
```


```{r echo = FALSE}
pkg_ver(c("lubridate", "stringr"))
```



Date values can be represented in tables as numbers or characters. But to be properly interpreted by R as dates, date values should be converted to an R **date** object class or a `POSIXct`/`POSIXt` object class. R provides many facilities to convert and manipulate dates and times, but a package called `lubridate` makes working with dates/times much easier.

&nbsp;&nbsp;

# Creating date/time objects
&nbsp;&nbsp;

## From complete date strings

You can convert many representations of date and time to date objects. For example, let's create a vector of dates represented as month/day/year character strings,

```{r}
x <- c("06/23/2013", "06/30/2013", "07/12/2014")
class(x)
```

At this point, R treats the vector `x` as characters. To force R to interpret these as dates, use `lubridate`'s `mdy` function. `mdy` will convert date strings where the date elements are ordered as month, day and year.

```{r}
library(lubridate)
x.date <- mdy(x)
class(x.date)
```

If you need to specify the time zone, add the parameter `tz=`. For example, to specify *Eastern Standard Time*, type:

```{r}
x.date <- mdy(x, tz="EST")
x.date
```

Note that using the `mode` or `typeof` functions will not help us determine if the object is an R *date* object. This is because a date is stored as a `numeric` (double) internally. Use the `class` function instead as shown in the above code chunk. 

The `mdy` function can read in date formats that use different delimiters so that `mdy("06/23/2013")`, `mdy("06-23-2013")` and `mdy("06.23.2013")` are parsed exactly the same so long as the order remains month/day/year.

For different month/day/year arrangements, other `lubridate` functions need to be used:

Functions  |  Date Format
-----------|--------------
`dmy()`    | day/month/year
`ymd()`    | year/month/day
`ydm()`    | year/day/month

If your data contains both date *and* time in a "month/day/year hour:minutes:seconds" format use the `mdy_hms` function.

```{r}
x <- c("06/23/2013 03:45:23", "07/30/2013 14:23:00", "08/12/2014 18:01:59")
x.date <- mdy_hms(x, tz="EST")
x.date
```

The characters `_h`, `_hm` or `_hms` can be appended to any of the four date function names described earlier to accommodate time formats. A few examples follow:

```{r}
mdy_h("6/23/2013 3", tz="EST") 
dmy_hm("23/6/2013 3:15", tz="EST")
ymd_hms("2013/06/23 3:15:7", tz="EST")
```

Note that adding a time element to the date object will  create `POSIXct` and `POSIXt` object classes instead of `Date` object classes.

```{r}
class(x.date)
```

Also, if a timezone is not explicitly defined for a time based date, the function assigns `UTC` ( Universal Coordinated Time).

```{r}
dmy_hm("23/6/2013 3:15")
```

## Setting and modifying timezones

R does not maintain its own list of timezone names, instead, it relies on the operating system's naming convention. To list the supported timezone names for your particular R environment, type:

```{r eval=FALSE}
OlsonNames()
```

For example, to select Daylight Savings Time type `tz = "EST5EDT"`.

```{r}
x.date <- mdy_hms(x, tz="EST5EDT")
x.date
```


```{r}
class(x.date)
```

If you need to convert the day/time to another timezone, use `lubridate`'s `with_tz()` function. For example, to convert `x.date` from it's current `EST5DST` timezone to the `US/Alaska` time zone, type:

```{r}
with_tz(x.date, tzone = "US/Alaska") 
```

Note that the `with_tz` function will change the timestamp to reflect the new time zone. If you simply want to change the time zone definition and not the timestamp, use the `tz()` function.

```{r}
tz(x.date) <- "US/Alaska"
x.date
```

## From separate date elements

If your data table splits the date elements into separate vector objects or columns, use the `paste` function to combine the elements into a single date string before passing it to one of the `lubridate` functions. Let's look at an example:

```{r}
dat1 <- read.csv("http://mgimond.github.io/ES218/Data/CO2.csv")
head(dat1)
```

The CO2 dataset has the date split across two columns: `Year` and `Month` (both stored as integers). You can combine the columns into a character string using the `paste` function. For example, if we want to create a "Year-Month" string as in `1959-10`, we could type:

```{r eval=FALSE}
paste(dat1$Year,dat1$Month, sep="-")
```

The above example uses three arguments: the two objects that are pasted together (i.e. `Year` and `Month`) and the `sep="-"` parameter which fills the gap between both objects with a dash `-` (by default, `paste` would have added spaces thus creating strings in the form of `1959 10`).

`lubridate` does not have a function along the lines of `ym` to convert just the year-month strings, this requires that we add an artificial day of the month to the string. We'll choose to add the 15^th^ day of the month as in 

```{r eval=FALSE}
paste(dat1$Year, dat1$Month, "15", sep="-")
```

And finally, we'll add a new column called `Date` to the `dat` object, and fill that column with the newly created date string wrapped with the `ymd` function:

```{r}
dat1$Date <- ymd( paste(dat1$Year, dat1$Month, "15", sep="-") )
head(dat1)
```

The `sep="-"` option is not needed with the lubridate function so the last piece of code could have been written as:

```{r}
dat1$Date <- ymd( paste(dat1$Year, dat1$Month, "15") )
```

To confirm that the `Date` column is indeed formatted as a date object type:

```{r}
str(dat1)
```

or you could type,

```{r}
class(dat1$Date)
```

Since we did not add a timezone or a time component to the date object the `Date` column was assigned a `Date` class as opposed to the `POSIX...` class. 

## Padding time values

The lubridate functions may expect the time values to consist of a specific number of characters if a delimiter such as `:` is not present to split the time elements. For example, the following will not generate a valide date/time object: 

```{r}   
hrmin <- 712           # Time 7:12
date  <- "2018/03/17"  # Date 
ymd_hm(paste(date, hrmin))
```

One solution is to *pad* the time element with 0's to complete a four character vector (or a six character vector if seconds are part of the time element). We can use the `str_pad` function from the `stringr` package to pad the time object (the `stringr` package is covered [in another tutorial](https://mgimond.github.io/ES218/Week03d.html)).

```{r}
library(stringr)
ymd_hm(paste(date, str_pad(hrmin, width=4, pad="0")))
```



# Extracting date information

If you want to extract the day of the week from a date vector, use the `wday` function. 

```{r}
wday(x.date) 
```

If you want the day of the week displayed as its three letter designation, add the `label=TRUE` parameter.

```{r}
wday(x.date, label=TRUE) 
```

You'll note that the function returns a `factor` object with seven levels--one for each day of the week (`r levels(lubridate::wday(x.date, label = TRUE))`)--as well as the level hierarchy which will dictate the order in which values will be displayed if grouped by this factor. The levels are not necessarily reflected in the vector elements (only `r unique(lubridate::wday(x.date, label=TRUE))` are present), but the levels are there if we were ever to add addition day elements to this vector.

The following table lists functions that extract different elements of a date object.

Functions        |  Extracted element
-----------------|--------------
`hour()`         | Hour of the day
`minute()`       | Minute of the hour
`day()`          | Day of the month
`yday()`         | Day of the year
`decimal_date()` | Decimal year
`month()`        | Month of the year
`year()`         | Year
`tz()`           | Time zone

# Operating on dates

You can apply certain operations to dates as you would to numbers. For example, to list the number of days between the first and third elements of the vector `x.date` type the following:

```{r}
(x.date[3] - x.date[1]) / ddays()
```

To get the number of weeks between both dates:
```{r}
(x.date[3] - x.date[1]) / dweeks()
```

Likewise, you can get the number of minutes between dates by dividing by `dminutes()` and the number of years by dividing by `dyears()`. 

You can also apply Boolean operations on dates. For example, to find which date element in `x.date` falls between the 11^th^ and 24^th^ day of any month, type:

```{r}
(mday(x.date) > 11) & (mday(x.date) < 24)
```

If you want the command to return just the dates that satisfy this query, pass the Boolean operation as an index to the `x.date` vector:

```{r}
x.date[ (mday(x.date) > 11) & (mday(x.date) < 24) ]
```

# Formatting date objects

You can create a character vector from a date object. This is useful if you want to annotate  plots with dates or include date values in reports. For example, to convert the date object `x.date` to a "Month_name Year" character format, type the following:

```{r}
as.character(x.date, format="%B %Y")
```

The `format=` parameter accepts many different date/time format codes listed in the following table (note the case!).

Format codes  |  Description             | Example
--------------|--------------------------|------------------------------------------
`%a`          | Abbreviated weekday name | `r as.character(x.date, format="%a")`
`%A`          | Full weekday name        | `r as.character(x.date, format="%A")`
`%m`          | Month as decimal number  | `r as.character(x.date, format="%m")`
`%b`          | Abbreviated month name   | `r as.character(x.date, format="%b")`
`%B`          | Full month name          | `r as.character(x.date, format="%B")`
`%c`          | Date and time, locale-specific |  `r as.character(x.date, format="%c")`
`%d`          | Day of the month as decimal number | `r as.character(x.date, format="%d")`
`%H`          | Hours as decimal number (00 to 23) | `r as.character(x.date, format="%H")`
`%I`          | Hours as decimal number (01 to 12) | `r as.character(x.date, format="%I")`
`%p`          | AM/PM indicator in the locale | `r as.character(x.date, format="%p")`
`%j`          | Day of year as decimal number | `r as.character(x.date, format="%j")`
`%M`          | Minute as decimal number (00 to 59) | `r as.character(x.date, format="%M")`
`%S`          | Second as decimal number | `r as.character(x.date, format="%S")`
`%U`          | Week of the year starting on the first Sunday | `r as.character(x.date, format="%U")`
`%W`          | Week of the year starting on the first Monday | `r as.character(x.date, format="%W")`
`%w`          | Weekday as decimal number (Sunday = 0) | `r as.character(x.date, format="%w")`
`%x`          | Date (locale-specific)  | `r as.character(x.date, format="%x")`
`%X`          | Time (locale-specific)  | `r as.character(x.date, format="%X")`
`%Y`          | 4-digit year            | `r as.character(x.date, format="%Y")`
`%y`          | 2-digit year            | `r as.character(x.date, format="%y")`
`%Z`          | Abbreviated time zone   | `r as.character(x.date, format="%Z")`
`%z`          | Time zone               | `r as.character(x.date, format="%z")`
