---
title: "Fine-tuning re-expressions"
output:
  html_document:
    toc: yes
    toc_depth: 3
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  comment = "",
  message = FALSE,
  tidy = FALSE,
#  dev=c('CairoPNG', 'CairoPDF'),
  type = "cairo",
  cache = FALSE,
  warning = FALSE,
  encoding = "UTF-8"
)
```

# Fine tuning re-expression

A lot can be learned by plotting the mid-summary values on a graph. Unlike the boxplot which gives us just one mid-summary value (i.e. the median), the letter values give us many mid-summary values.

The mids drop steadily (at least for the first three letter values) indicating skeweness towards lower values for the mid three quarters of the batch. If all mid-summary values were equal, then we would have a perfectly balanced batch. If the mid-summary


```{r}
library(dplyr)
library(ggplot2)
source("http://mgimond.github.io/ES218/es218.R")

df <- read.csv("http://mgimond.github.io/ES218/Data/Food_web.csv")

ggplot(df) +aes(y=bc(mean.length,0), x=dimension) +geom_boxplot()

df %>% group_by(dimension) %>% 
  do(as.data.frame( lsum(.$mean.length, l=6) )) %>%
  ggplot(.) + aes(x=letter, y=bc(mid,-1)) + geom_point() + facet_grid(.~dimension)
```
