---
title: "Spread-level plots"
output:
  html_document:
    toc: yes
    toc_depth: 3
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  comment = "",
  message = FALSE,
  tidy = FALSE,
#  dev=c('CairoPNG', 'CairoPDF'),
  type = "cairo",
  cache = FALSE,
  warning = FALSE,
  encoding = "UTF-8"
)
```

# Introduction

Some batches of data may show a systematic change in spread vs. location; a good example is the stereogram fusion time data in Cleveland's book where an increase in location translates into an increase in spread. Such dependency is often undesirable (e.g. in an ANOVA for instance) and preferably removed in an analysis. A plot well suited for visualizing this dependency is the **spread-level** plot, **s-l** (or **spread-location** plot as Cleveland calls it).

# Constructing the s-l plot

The plot compares a measure of the spread's residuals to the location (usually the median) for each batch of data. The spread is usually distilled down to its residuals (what remains after subtracting each batch value by the batch median) then transformed by taking the square root of its absolute value. Taking the absolute value collapses the spread about the median to prevent symmetric spreads from masking any systematic change in spread vs. location. The square root helps reduce the strong skew usually present in the residuals.

The following block of code recreates the s-l plot in figure 2.25 from Cleveland's book.

```{r, fig.height=4,fig.width=3.5}
library(dplyr)

df <- read.csv("http://mgimond.github.io/ES218/Data/fusion.csv")

# Create two new columns: group median and group residual
df1 <- df %>%
       group_by(nv.vv)   %>%
       mutate( med = median(time),
               res = sqrt( abs( time - med)))

# Compute the coordinate values needed to connect residual medians
sl.line <- df1 %>%
           group_by(nv.vv)  %>%
           summarise(location = median(time),
                     res.med  = median(res))

# Generate the s-l plot
plot( res ~ jitter( med, am=0), dat=df1, pch=20, col=rgb(0,0,0,0.3),
      xlab = "Median time (jittered)",
      ylab = "Sqrt of absolute residual time")

# Connect the residual medians
lines( sl.line$location, sl.line$res.med, col="red")

# Add group labels to batches
mtext(sl.line$nv.vv, side=3, at=sl.line$location)

```

The red line in the plot helps identify the type of relationship between spread and location. If the line slopes upward, there is an increasing spread as a function of increasing location; if it slopes downward, there is a decreasing spread as a function of increasing location; and if the slope is flat, there is no change in spread as a function of location.

In the data fusion example, there is an apparent increase in spread as a function of location.


# Other examples

## Cleveland's food web data

On page 60, Cleveland compares spread vs location for three different batches of values: two, three and mixed dimensional ecosystems. The values represent the mean chain length of a web (defined by the  number of hierarchical prey-predator links in the web) for the three different ecosystem types. We will modify the chunk of code used in generating the s-l plot in the last section to sort the ecosystem dimensions by group median (i.e. we want the batches to be ordered from smallest median to largest median along the x-axis); this is done in the last three lines of the piping operation. We would normally not need to sort numeric values (recall that the x-axis represents the median of each batch, hence a number), but we do so to facilitate the plotting of the connecting lines.

```{r, fig.height=4,fig.width=3.5}
df <- read.csv("http://mgimond.github.io/ES218/Data/Food_web.csv")

# Create two new columns: group median and group residual
df1 <- df %>%
       group_by(dimension)                            %>%
       mutate( med = median(mean.length),
               res = sqrt( abs( mean.length - med)))  %>%
       ungroup()                                      %>%
       arrange(med)                                   %>%
       mutate( dimension = factor(dimension, levels=unique(dimension)))


# Compute the coordinate values needed to connect residual medians
sl.line <- df1 %>%
           group_by(dimension)  %>%
           summarise(location = median(mean.length),
                     res.med  = median(res))

# Generate the s-l plot
plot( res ~ jitter( med, am=0), dat=df1, pch=20, col=rgb(0,0,0,0.3),
      xlab = "Median time (jittered)",
      ylab = "Sqrt of absolute residual time")

# Connect the residual medians
lines( sl.line$location, sl.line$res.med, col="red")

# Add group labels to batches
mtext(sl.line$dimension, side=3, at=sl.line$location)
```

A monotonic spread is apparent in this dataset too, i.e. as the median chain length increases, so does the spread.

## Buoy data

In this next example, we will compare water temperature spreads with median temperatures across 12 months using Gulf of Maine buoy data. Since we have many categories to work with, we will do without the jitter and stack the points within each month.

```{r, fig.height=4,fig.width=8}
df <- read.table("http://mgimond.github.io/ES218/Data/buoy_44005_2012.dat", header=T)

# Create two new columns: group median and group residual
df1 <- df %>%
       group_by(MM)                            %>%
       mutate( med = median(WTMP),
               res = sqrt( abs( WTMP - med)))  %>%
       ungroup()                                      %>%
       arrange(med)                                   %>%
       mutate( MM = factor(MM, levels=unique(MM)))


# Compute the coordinate values needed to connect residual medians
sl.line <- df1 %>%
           group_by(MM)  %>%
           summarise(location = median(WTMP),
                     res.med  = median(res))

# Generate the s-l plot
plot( res ~ med, dat=df1, pch=20, col=rgb(0,0,0,0.1),
      xlab = "Median water temperature (jittered)",
      ylab = "Sqrt of absolute water temperature")

# Connect the residual medians
lines( sl.line$location, sl.line$res.med, col="red")

# Add group labels to batches
mtext(sl.line$MM, side=3, at=sl.line$location, cex=0.5)
```

Note that the discrete appearance of the values reflects the level of precision used to record  water temperature data (a tenth of a degree Celsius). Unlike our previous two examples, a systematic spread is not immediately apparent.

# Variations of the S-L plots

Another version of the S-L plot (and one that seems to be the most popular) pits the log of the "hinges" spread vs the log of the median. This approach only works for positive values (this may require that values be adjusted so that the minimum value be no less than or equal to 0).  This approach is appealing in that the slope of the best fit line can be used to come up with a power transformation. 

The hinges spread is similar to the interquartile range but because it is computed differently, it can take on slightly different values. The hinge is the half-way point between the most extreme value of a batch and the median. As such, there is a left hinge and a right hinge. Hinges (and other depths) are discussed in another section titled "Letter value summaries".

The aforementioned s-l plot can be computed in R as follows (we will use the buoy data as an example).

```{r fig.height=4, fig.width=4}
sl <- df %>% 
      group_by(MM)  %>%
      summarise (level  = log(median(WTMP)),
                 frth   = (floor( (n()+1)/2) + 1)/2,
                 lower  = sort(WTMP)[floor(frth)] ,
                 upper  = sort(WTMP)[ceiling(n() + 1 - frth)],
                 sprd   = log(upper - lower) )

plot(sprd ~ level, sl, pch=16, xlab="Median (log)", ylab="Spread (log)")
M <- lm(sprd ~ level, sl) 
abline( M, col="red")
text(sl$level,sl$sprd, labels=sl$MM, pos=4, cex=0.6)
```

Note how this plot differs from Cleveland's plot in that we are only displaying each month's median spread value and we are fitting a straight line to the points.

The slope suggests that there may be a monotonic increase in spread vs location, but a close examination of the regression line indicates that the slope may not be significantly different from 0.

```{r}
summary(M)
```

