---
title: "Relational and boolean operations"
editor_options: 
  chunk_output_type: console
---

```{r echo=FALSE}
source("libs/Common.R")
options(width = 80)
```

You've already been exposed to a few examples of relational and boolean operations in earlier tutorials. A formal exploration of these techniques follow.

# Relational operations

Relational operations play an important role in data manipulation. Anytime you subset a dataset based on one or more criterion, you are making use of a relational operation. The relational operators (also known as *logical binary operators*) include `==`, `!=`, `<`, `<=`, `>` and `>=`. The output of a condition is a logical vector `TRUE` or `FALSE`.  
<br>  

+-----------------------+--------+------------------+
| Relational operator   | Syntax | Example          |
+=======================+========+==================+
| Exact equality        | `==`   |  3 == 4 -> FALSE |
+-----------------------+--------+------------------+
| Exact inequality      | `!=`   |  3 != 4 -> TRUE  |
+-----------------------+--------+------------------+
| Less than             | `<`    |  3 < 4  -> TRUE  |
+-----------------------+--------+------------------+
| Less than or equal    | `<=`   |  4 <= 4 -> TRUE  |
+-----------------------+--------+------------------+
| Greater than          | `>`    |  3 > 4  -> FALSE |
+-----------------------+--------+------------------+
| Greater than or equal | `>=`   |  4 >= 4 -> TRUE  |
+-----------------------+--------+------------------+

# Boolean operations

Boolean operations can be used to piece together multiple evaluations.

R has three boolean operators: The **AND** operator, `&`; The **NOT** operator, `!`; And the **OR** operator, `|`.

The `&` operator requires that the conditions on both sides of the boolean operator be satisfied. You would normally use this operator when addressing a question along the lines of *"`x` must be satisfied AND `y` must be satisfied"*.

The `|` operator requires that at least one condition be met on either side of the boolean operator. You would normally use this operator when addressing a question along the lines of "`x` must be satisfied OR `y` must be satisfied". Note that the output will also be TRUE if *both* conditions are met.

The `!` operator is a *negation* operator. It will reverse the outcome of a condition. It can be interpreted as *"I do NOT want `x` to be true"*. So if the outcome of an expression is `TRUE`, preceding that expression with `!` will reverse the outcome to `FALSE` and vice-versa.  
<br> 
  
+----------+----------+------------------------+--------------+
| Boolean  | Syntax   | Example                | Outcome      |
| operator |          |                        |              |
+==========+==========+========================+==============+
| AND      | `&`      | 4 == 3 `&` 1 == 1 <br> | FALSE  <br>  |
|          |          | 4 == 4 `&` 1 == 1      | TRUE         |
+----------+----------+------------------------+--------------+
| OR       | `|`      | 4 == 4 `|` 1 == 1 <br> | TRUE  <br>   |
|          |          | 4 == 3 `|` 1 == 1 <br> | TRUE  <br>   |
|          |          | 4 == 3 `|` 1 == 2      | FALSE        |
+----------+----------+------------------------+--------------+
| NOT      | `!`      | `!`(4 == 3) <br>       | TRUE  <br>   |
|          |          | `!`(4 == 4)            | FALSE        |
+----------+----------+------------------------+--------------+

The following table breaks down all possible Boolean outcomes where `T` = `TRUE` and `F` = `FALSE`:

+-------------------+------------+
| Boolean operation | Outcome    |
+===================+============+
| T `&` T           | TRUE       |
+-------------------+------------+
| T `&` F           | FALSE      |
+-------------------+------------+
| F `&` F           | FALSE      |
+-------------------+------------+
| T `|` T           | TRUE       |
+-------------------+------------+
| T `|` F           | TRUE       |
+-------------------+------------+
| F `|` F           | FALSE      |
+-------------------+------------+
| `!`T              | FALSE      |
+-------------------+------------+
| `!`F              | TRUE       |
+-------------------+------------+

If the input values to a boolean operation are numeric vectors and not logical vectors, the numeric values will be interpreted as `FALSE` if zero and `TRUE` otherwise. For example:

```{r}
1 & 2
1 & 0
```

## A word of caution

Note that the operation `a == (3 | 4)` is **not** the same as `(a == 3) | (a == 4)`. The former will return `FALSE` whereas the latter will return `TRUE` if `a = 3`. This is because the Boolean operator evaluates both sides of its expression as separate **logical** outcomes (i.e. `T` and `F` values). In the latter case, the Boolean expression is asking *"is `a` equal to `3` OR is `a` equal to `4`"*. Since one of the conditions is true, the expression ends up evaluating `TRUE | FALSE` which returns `TRUE` (see above table).

```{r}
a <- 3
b <- 4
(a == 3) | (a == 4)
```

In the former expression, the boolean operator `|` is evaluating `3` OR `4` on its right-hand side. As mentioned in the previous section, logical values take on a value of `0` for FALSE and any non-zero value for TRUE, so when evaluating `3 | 4`, it's really seeing `TRUE | TRUE` which, according to the aforementioned table will output `TRUE`.

```{r}
3 | 4
```

So in the end, the expression `a == (3 | 4)` is really evaluating the condition `a == TRUE` which returns false (since 3 is not equal to the logical value `TRUE`).

```{r}
a == (3 | 4)
```

# Comparing multidimensional objects

The relational operators are used to compare single elements (i.e. one element at a time). If you want to compare two objects as a whole (e.g. multi-element vectors or data frames), use the `identical()` function. For example:

```{r}
a <- c(1, 5, 6, 10)
b <- c(1, 5, 6)
identical(a, a)
identical(a, b)
identical(mtcars, mtcars)
```

Notice that `identical` returns a single logical vector, regardless the input object's dimensions.

Note that the data structure must match as well as its element values. For example, if `d` is a list and `a` is an atomic vector, the output of `identical` will be false even if the internal values match.

```{r}
d <- list( c(1, 5, 6, 10) )
identical(a, d)
```

If we convert `d` from a list to an atomic vector using the `unlist` function (thus matching data structures), we get:

```{r}
identical(a, unlist(d))
```

# The match operator `%in%` 

The match operator `%in%` compares two sets of vectors and assesses if an element on the left-hand side of `%in%` matches any of the elements on the right-hand side of the operator. For each element in the left-hand vector, R returns `TRUE` if the value is present in any of the right-hand side elements or `FALSE` if  not. 

For example, given the following vectors:

```{r}
v1 <- c( "a", "b", "cd", "fe")
v2 <- c( "b", "e")
```

find the elements in `v1` that match any of the values in `v2`. 

```{r}
v1 %in% v2
```

The function checks whether each element in `v1` has a matching value in `v2`. For example, element `"a"` in `v1` is compared to elements `"b"` and `"e"` in `v2`. No matches are found and a `FALSE` is returned. The next element in `v1`, `"b"`, is compared to both elements in `v2`. This time, there is a match (`v2` has an element `"b"`) and  `TRUE` is returned. This process is repeated for all elements in `v1`.

The logical vector output has the same length as the input vector `v1` (four in this example).

If we swap the vector objects, we get a two element logical vector since we are now comparing each element in `v2` to any matching elements in `v1`.

```{r}
v2 %in% v1
```

<center>
<video width="620"  controls  style="float:center;">
  <source src="./Videos/Matching_operator.mp4" type="video/mp4">
</video>
</center>

# Checking if a value is `NA`

When assessing if a value is equal to `NA` the following evaluation may behave unexpectedly.

```{r}
a <- c (3, 67, 4, NA, 10)
a == NA
```

The output is not a logical data type we would expect from an evaluation. Instead, you must make use of the `is.na()` function:

```{r}
is.na(a)
```

As another example, if we want to keep all rows in dataframe `d` where `z` = `NA`, we would type:

```{r}
d <- data.frame(x = c(1,4,2,5,2,3,NA), 
                y = c(3,2,5,3,8,1,1), 
                z = c(NA,NA,4,9,7,8,3))

d[ is.na(d$z), ]
```

You can, of course, use the `!` operator to reverse the evaluation and *omit* all rows where `z` = `NA`,

```{r}
d[ !is.na(d$z), ]
```