---
title: "Customizing colors in R"
---

```{r echo=FALSE}
source("libs/Common.R")
```

```{r echo = FALSE}
R_ver(c)
```


```{css, echo=FALSE}
.scroll {
   overflow-y: scroll;
    height: 100px;
}
```

----

R has a large number of built-in colors that can be accessed by their names--`r length(colors(distinct = FALSE))` as of this writing . But many of these color names share the same RGB colors thus reducing the set of *unique* colors to `r length(colors(distinct = TRUE))`.

To see the list of distinct colors type:

```{r class.output="scroll", R.options = list(width = 100)}
colors(distinct = TRUE)
```

Built-in colors can be accessed by their names. For example, to fill the boxplot with a `"bisque"` color, type:

```{r fig.width = 3, fig.height = 1.5, echo = 2}
OP <- par(mar = c(0,0,0,1), bty="n")
boxplot(mtcars$mpg, axes=FALSE, horizontal = TRUE, col = "bisque")
par(OP)
```

## Color palettes

R has built-in color palettes that can save you from having to manually construct a palette of colors. 


Prior to R version `4.1`, the default color palette is defined by the following colors:

```{r eval = FALSE}
palette()
```

```{r echo = FALSE}
palette("R3"); palette(); palette("R4") # Revert back to modern palette
```

With R version `4.1` and later, the default color palette is defined by the following colors:

```{r}
palette()
```

Note that the latter set of colors are defined in hexadecimal color code. The following plot shows their colors.

```{r fig.width=9, fig.height=1, echo=FALSE}
OP <- par(mar=c(1,0,0,0) , mgp=c(0,-2,0) )
col.lst <- palette()
plot(1:length(col.lst), rep(1,length(col.lst)), axes = FALSE, pch=16, cex=2, col=col.lst, xlab=NA, ylab=NA,
     xlim = c(0, 11))
axis(1, at=1:length(col.lst), labels = sprintf("%s", col.lst), col="white", cex=0.6, padj=-3, cex.axis=0.8)
par(OP)
```

R version `4.1` and later offers 16 built-in color palettes (earlier versions of R only offer one). The palette names (and their colors) are shown next.

```{r echo = FALSE, fig.width = 8, fig.height=5, message=FALSE}
library(ggplot2)
cols <- sapply(palette.pals(), palette.colors, n= NULL, recycle = FALSE)
cols.df <- data.frame(ID = rep(names(cols), sapply(cols, length)),
           Obs = unlist(cols))
cols.df$ID <- factor(cols.df$ID,names(cols) )
ggplot(cols.df) + aes(x=ID) + 
  geom_dotplot(fill=cols.df$Obs,col=cols.df$Obs) + coord_flip() +
  theme_classic() +
  theme(axis.line = element_blank(), 
        axis.title = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks = element_blank())
```

Palette `R4` is the default color palette for R version `4.1` and later. Palette `R3` is the color palette used in earlier versions of R.

With R version `4.1` and later, you can access anyone of these palettes using the `palette` command. For example, if you want to use the `Classic Tableau` color scheme in a plot, precede the plot command with the call to `palette("Classic Tableau")`.

```{r fig.width = 3, fig.height= 3, echo=2:3}
OP <- par(mar = c(2,2,1,0))
palette("Classic Tableau")
boxplot(log(decrease) ~ treatment, data = OrchardSprays,
        col = 1:length(levels(OrchardSprays$treatment)))
par(OP)
```

The argument `1:length(levels(OrchardSprays$treatment))` is generating the sequence of integer values reflecting the number of unique levels. Each number designates a color index in the color palette--one for each boxplot generated in the plot. For example, the first boxplot is assigned a color index of `1` or:

```{r}
palette()[1]
```
The second boxplot is assigned the color index of `2`, or:

```{r}
palette()[2]
```
We could have written the boxplot command as follows:

```{r eval = FALSE}
boxplot(log(decrease) ~ treatment, data = OrchardSprays,
        col = c(palette()[1], palette()[2], palette()[3], palette()[4],
                palette()[5], palette()[6], palette()[7], palette()[8]))
```

Or, we could have explicitly defined the colors using their hexadecimal representation:

```{r eval = FALSE}
boxplot(log(decrease) ~ treatment, data = OrchardSprays,
        col = c("#1F77B4", "#FF7F0E", "#2CA02C", "#D62728", 
                "#9467BD", "#8C564B", "#E377C2", "gray50"))
```

Note that the above boxplots only make use of `Classic Tableau`'s first 8 color swatches (the palette has `10` color swatches). Note that if there are fewer colors passed to the `col =` argument than there are grouping variables, the colors get recycled. For example, if we pass just three colors to the `col =` argument, we get:

```{r fig.width = 3, fig.height= 3, echo=2}
OP <- par(mar = c(2,2,1,0))
boxplot(log(decrease) ~ treatment, data = OrchardSprays,
        col = c("#1F77B4", "#FF7F0E", "#2CA02C"))
par(OP)
```

Note how the blue, orange and green colors get recycled.

To revert back to the default color palette, precede the plot command with another call to `palette()` using the `"R4"` argument.:

```{r fig.width = 3, fig.height= 3, echo=2:3}
OP <- par(mar = c(2,2,1,0))
palette("R4")
boxplot(log(decrease) ~ treatment, data = OrchardSprays,
        col = 1:length(levels(OrchardSprays$treatment)))
par(OP)
```



## RGB designation

Colors can also be defined by the computer display's primary colors; **red**, **green** and **blue** (**RGB** for short). One can use the `rgb()` function to define the levels of each primary color from a scale of 0 to 1 with one being the color's maximum intensity. For example, if you want to generate a pure blue color, type:

```{r}
rgb(0,0,1)
```

The three `rgb()` values are assigned to the *red*, *green* and *blue* colors respectively. In this example, red and green are assigned the values `0` and blue is assigned the value of `1`.
The function returns a hexadecimal value, `r rgb(0,0,1)`, which R can convert to a color. For example to plot blue points, type:

```{r fig.height=3, fig.width=3, echo=2}
OP <- par(mar=c(4,4,0,0))
plot( speed ~ dist, cars, pch = 16, col = rgb(0, 0, 1))
par(OP)
```

This is the same as typing:

```{r eval = FALSE}
plot( speed ~ dist, cars, pch = 16, col = "#0000FF")
```

### Applying transparency

A fourth parameter can be passed to the `rgb()` function: the opaqueness value `alpha`. This is useful when many overlapping points are displayed on a plot. `alpha` is defined in an interval from `0` to `1` with `1` being completely opaque and 0 being completely transparent. In the following example, we apply a blue color to the dots and assign them a 80% transparency (opaqueness = 0.2):

```{r fig.height=3, fig.width=3, echo=2}
OP <- par(mar=c(4,4,0,0))
plot( Sepal.Width ~ Sepal.Length, iris , pch=16, col=rgb(0, 0, 1, 0.2))
par(OP)
```
