---
title: "Week 3: Manipulating data tables with dplyr"
---

```{r echo=FALSE}
source("libs/Common.R")
```


<div style="color:#ff7535; background-color:#fff0ee ;	border-left-style: solid">
This tutorial makes use of the following R package(s): **`dplyr`**
</div>
 


The data file  *FAO_grains_NA.csv* will be used in this exercise. This dataset consists of grain yield and harvest year by north american country. The dataset was downloaded from http://faostat3.fao.org/ in June of 2014. 

Run the following line to load the FAO data file into your current R session.

```{r}
dat <- read.csv("http://mgimond.github.io/ES218/Data/FAO_grains_NA.csv", header=TRUE)
```


# The `dplyr` basics

The basic set of R tools can accomplish many data table queries, but the syntax can be overwhelming and verbose. The package `dplyr` offers some nifty and simple querying functions as shown in the next subsections. `dplyr`'s data manipulation functions are summarized in the following table:

`dplyr` function   | Description
-------------------|-------------
`filter()`         | Subset by row values
`arrange()`        | Sort rows by column values
`select()`         | Subset columns
`mutate()`         | Add columns
`summarise()`      | Summarize columns

Make sure to load the `dplyr` package before proceeding with the following examples. 

```{r, message=FALSE}
library(dplyr)
```

## Subset by rows: `filter`

Tables can be subsetted by rows based on column values. For example, we may wish to grab all rows associated with `Oats`: 

```{r}
dat.query1 <- filter(dat, Crop == "Oats")
summary(dat.query1)
```

Note that R is case sensitive, so make sure that you respect each letter's case (i.e. upper or lower).

We can expand our query by including both `Oats`, `Buckwheat` and limiting the country to `Canada`.

```{r}
dat.query2 <- filter(dat, Crop == "Oats" | Crop == "Buckwheat", 
                          Country == "Canada")
summary(dat.query2)
```

The character `|` is the Boolean operator OR. So in our example, the query can be read as "... crop equals oats OR crop equals buckwheat". Had we used the AND operator, `&`, instead as in `Crop == "Oats" & Crop == "Buckwheat"` the output would have returned zero rows since a `Crop` value cannot be both `Oats` AND `Buckwheat`.

We can expand this query by limiting our output to the years 2005 to 2010

```{r}
library(dplyr)
dat.query3 <- filter(dat, Crop == "Oats" | Crop == "Buckwheat", 
                          Country == "Canada", 
                          Year >= 2005 & Year <= 2010)
summary(dat.query3)
```

Note the use of the AND Boolean operator (`&`) instead of the OR operator (`|`) for the `Year` query. We want the `Year` value to satisfy two criteria simultaneously: *greater than or equal to 2005* **AND** *less than or equal to 2010*. Had we used the `|` operator, R would have returned all years since all year values satisfy at least one of the two criteria.

## Sort rows by column value: `arrange`

You can sort a table based on a column's values. For example, to sort `dat` by crop name type:

```{r}
dat.sort1 <- arrange(dat, Crop)
head(dat.sort1)
tail(dat.sort1)
```

By default, `arrange` sorts by ascending order. To sort by descending order, wrap the column name with the function `desc()`. For example, to sort the table by `Crop` in ascending order then by `Year` in descending order, type:

```{r}
dat.sort2 <- arrange(dat, Crop, desc(Year))
head(dat.sort2)
tail(dat.sort2)
```

## Subset by column: `select`

You can grab a subset of the table by column(s). To extract the columns `Crop`, `Year` and `Value`, type:

```{r}
dat.subcol <- select(dat, Crop, Year, Value)
head(dat.subcol)
```

## Adding columns: `mutate`

You can add columns (and compute their values) using the `mutate` function. For example, to add a column `Ctr_abbr` and assign it the abbreviated values `CAN` for Canada and `USA` for the United States of America based on the values in column `Country` type:

```{r}
dat.extended <- mutate(dat, Ctr_abbr = ifelse(Country == "Canada", "CAN", "USA"))
head(dat.extended)
tail(dat.extended)
```

Here, we make use of an embedded function, `ifelse`, which performs a conditional operation: if the `Country` value is `Canada` return `CAN` if not, return `USA`.

Note that if you wish to *rename* a column, you can use the `rename()` function instead of `mutate`.

## Summarize columns: `summarise`

You can summarize (or "collapse") one or more columns using the `summarise` function. For instance, to get the minimum and maximum years from the `Year` column, type:

```{r}
summarise(dat, yr_min = min(Year), yr_max=max(Year))
```

## Splitting data operations by group: `group_by`

The `group_by` function will split any operations applied to the dataframe into groups defined by one or more columns. For example, if we wanted to get the minimum and maximum years from the `Year` column for which crop data are available *by crop type*, we would type the following:

```{r}
dat.grp <- group_by(dat, Crop)
summarise(dat.grp, yr_min = min(Year), yr_max=max(Year))
```

The `dat.grp` object is a special data structure which differs from that of `dat` in that it explicitly defines the groups for which any subsequent operations (like `summarise`) are to be split across. 

## `dplyr`'s output data structure

Many of `dplyr`'s functions will add one or more classes to a data table. For example, if we stored the summary output from the last chunk of code to an object called `dat.sum`, we'd discover that this object is associated with three classes.

```{r}
dat.sum <- summarise(dat.grp, yr_min = min(Year), yr_max=max(Year))
class(dat.sum)
```

This may prove to be problematic when using this table in none `dplyr` functions. To remedy this, you can force the `dat.sum` object to a standalone `dataframe` as follows:

```{r}
dat.df <- data.frame(dat.sum)
class(dat.df)
```

# Combining table manipulation functions using `%>%`

In most cases, you will find yourself wanting to combine several of `dplyr`'s data manipulation functions. For example, to summarize the total yield of oats by country, you could type the following:

```{r}
dat.yield <- filter(dat, Information == "Yield (Hg/Ha)", Crop == "Oats")
dat.grp   <- group_by(dat.yield, Country)
dat.sum   <- summarise(dat.grp, mean_yield = mean(Value))
data.frame(dat.sum)
```

The one downside to this approach of combining data table manipulation functions for complex queries is the use of several intermediate objects (e.g. `dat.yield`, `dat.grp`, `dat.sum`). This can make the workflow difficult to follow and clutter your R session with needless intermediate objects. 

Another approach to combining `dplyr` operations is to use the **piping operator** ,`%>%`, which daisy chains `dplyr` operations. So our previous workflow would look like:

```{r}
dat %>%
  filter(Information == "Yield (Hg/Ha)", Crop == "Oats") %>%
  group_by(Country) %>%
  summarise(mean_yield = mean(Value)) %>%
  data.frame()
```

The chunk of code can be read as *"... with the `dat` table, `filter` by ..., then `group_by` ...., then `summarise` by ..."* with the result from one operation being passed on to the next using the `%>%` operator. Note that the `filter`, `group_by` and `summarise` functions do not include the data table name making the chunk of code less cluttered and easier to read.

This chunk of code will output the results to the screen, if you want to save it to an object called `dat2`, replace the first line with `dat2 <- dat %>% ...`.

Here's a more complicated output whereby the median yield for each USA crop is summarized by year for the years spanning 2005 through 2010 and the output is sorted by crop name (in ascending order) and median yield value (in descending order).

```{r}
dat %>%
  filter(Information == "Yield (Hg/Ha)", Year >= 2005, Year <=2010, 
         Country=="United States of America") %>%
  group_by(Crop, Year) %>%
  summarise(median_yield = round(median(Value))) %>%
  arrange(Crop, desc(median_yield))
```

# More on the `group_by` function

The `group_by` function followed by the `summarise` function will generate a new table with as many rows as there are groups defined in the `group_by` function. But grouping variables can be used for other purposes where the output table contains the same number of rows as the input table. The following sub-sections highlight a few examples.

## Normalizing each value in a group by the group median

In this example, we are subtracting each value in a group by that group's median. This can be useful in identifying which year yields are higher than or lower than the median yield value within each crop group. We will concern ourselves with US yields only and sort the output by crop type. We'll save the output dataframe as `dat2`.

```{r}
dat2 <- dat %>% 
  filter(Information == "Yield (Hg/Ha)",
         Country == "United States of America") %>%
  select(Crop, Year, Value)                     %>%
  group_by(Crop)                                %>%
  mutate(NormYield = Value - median(Value))     %>%
  arrange(Crop)
```

Let's plot the normalized yields by year for Barley and add a `0` line representing the (normalized) central value.

```{r fig.width=5, fig.height=3,small.mar=TRUE}
plot( NormYield ~ Year, dat2[dat2$Crop == "Barley",] )
abline(h = 0, col="red")
```

The relative distribution of points does not change, but the values do (they are re-scaled) allowing us to compare values based on some localized (group) context. This technique will prove very useful later on in the course when EDA topics are explored.


